---
title: "compare_clustering_sim3_ten_chr"
author: "Xuan Chen"
date: "2026-02-28"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = T)
rm(list=ls())
library(tidyverse)
library(mclust)
library(umap)
library(Rtsne)
library(rstan)
library(magrittr)
knitr::opts_knit$set(root.dir = "C:/Users/67402/OneDrive - University of Miami/courses/research/HiC")
```


```{r}
source("Functions/Functions.R")
```

```{r}
lambda_list = c(0.01,0.03,0.05,0.1,0.2,0.3,0.4,0.5,0.7,1)
set.seed(123456)
n_rep = 10
resolution = 50*1e3
```


```{r}
load("Output/output_251026_sim_HBA_ten_chr_combine/muS_combined_plot_260209.RData")
```


# Functions

```{r}
# ----------------------------
# 0) Helpers: get truth labels
# ----------------------------
get_cell_type_label <- function(chr_id, true_by_chr, substr_n = 10) {
  # CHANGE: robust coercion to a single character key
  chr_key <- as.character(chr_id[1])

  obj <- true_by_chr[[chr_key]]
  if (is.null(obj)) stop("true_by_chr has no entry for chr = ", chr_key)

  lab <- colnames(obj)
  if (is.null(lab)) stop("true_by_chr[[", chr_key, "]] has NULL colnames()")

  substr(lab, 1, substr_n)
}

# -------------------------------------------------
# 1) Unified normalization (NO "0 -> 1" coercion)
# -------------------------------------------------
normalize_unified <- function(mat,
                              per_cell_scale = TRUE,
                              log1p_transform = TRUE,
                              zscore_features = TRUE) {
  mat2 <- mat

  if (per_cell_scale) {
    cs <- colSums(mat2)
    ok <- cs > 0
    if (any(ok)) {
      mat2[, ok] <- t(t(mat2[, ok]) / cs[ok])
    }
    # leave cs==0 columns as-is (all zeros), don't fake them
  }

  if (log1p_transform) {
    # valid for nonnegative continuous too
    mat2 <- log1p(mat2)
  }

  if (zscore_features) {
    mat2 <- t(scale(t(mat2)))
    mat2[is.na(mat2)] <- 0
  }

  mat2
}
```

```{r}
# -----------------------------------------
# 2) Embedding: UMAP or tSNE (cells in rows)
# -----------------------------------------
# --------------------------------------------
# CHANGE: new helper to drop zero-variance features
# --------------------------------------------
drop_constant_features <- function(mat, eps = 0) {   # mat is N x K (features x cells)
  # variance across cells for each feature (row)
  v <- apply(mat, 1, stats::var)
  keep <- is.finite(v) & (v > eps)
  mat2 <- mat[keep, , drop = FALSE]
  list(mat = mat2, keep = keep)
}

# --------------------------------------------
# CHANGE: embed_cells updated to drop constant columns + safer tSNE PCA scaling
# --------------------------------------------
embed_cells <- function(mat,
                        embed_method = c("umap", "tsne"),
                        n_neighbors = 10,
                        tsne_perplexity = NULL,
                        seed = 12345) {
  embed_method <- match.arg(embed_method)

  # mat: N x K (features x cells)
  # CHANGE: remove constant/zero-variance features BEFORE transposing
  dropped <- drop_constant_features(mat)
  mat_use <- dropped$mat

  # If everything becomes constant -> cannot embed
  if (nrow(mat_use) == 0) {
    return(NULL)  # caller handles NULL
  }

  X <- t(mat_use)  # K x N : cells x features

  set.seed(seed)
  if (embed_method == "umap") {
    cfg <- umap.defaults
    cfg$n_neighbors  <- n_neighbors
    cfg$random_state <- seed
    res <- umap(X, config = cfg)
    emb <- as.data.frame(res$layout)
    colnames(emb) <- c("X1", "X2")
    return(emb)
  } else {
    K <- nrow(X)
    if (is.null(tsne_perplexity)) {
      tsne_perplexity <- min(30, floor((K - 1) / 3))
      tsne_perplexity <- max(tsne_perplexity, 2)
    }

    # CHANGE: set pca_scale = FALSE to prevent "rescale constant column" error
    ts <- Rtsne(
      X,
      dims = 2,
      perplexity = tsne_perplexity,
      check_duplicates = FALSE,
      pca = TRUE,
      pca_scale = FALSE,   # <-- CHANGE
      verbose = FALSE
    )
    emb <- as.data.frame(ts$Y)
    colnames(emb) <- c("X1", "X2")
    return(emb)
  }
}

```


```{r}
# -------------------------------------------------------
# 3) One-shot ARI: normalize(optional) → embed → kmeans → ARI
#    IMPORTANT: ARI computed vs true cell_type_label (robust)
# -------------------------------------------------------
# --------------------------------------------
# CHANGE: compute_ari_one updated to handle emb==NULL
# --------------------------------------------
compute_ari_one <- function(mat,
                            true_labels,
                            embed_method = c("umap", "tsne"),
                            normalize = TRUE,
                            norm_args = list(per_cell_scale=TRUE, log1p_transform=TRUE, zscore_features=TRUE),
                            n_neighbors = 10,
                            tsne_perplexity = NULL,
                            nstart = 10,
                            seed = 12345) {

  embed_method <- match.arg(embed_method)

  # guard: no information
  if (all(mat == 0) || stats::sd(as.vector(mat)) == 0) return(NA_real_)

  mat_use <- mat
  if (normalize) {
    mat_use <- do.call(normalize_unified, c(list(mat = mat_use), norm_args))
  }

  emb <- embed_cells(
    mat_use,
    embed_method = embed_method,
    n_neighbors = n_neighbors,
    tsne_perplexity = tsne_perplexity,
    seed = seed
  )

  # CHANGE: if embedding failed (all features constant), return NA
  if (is.null(emb)) return(NA_real_)

  k <- length(unique(true_labels))
  set.seed(seed)
  km <- kmeans(emb, centers = k, nstart = nstart)

  adjustedRandIndex(km$cluster, as.integer(factor(true_labels)))
}

```

```{r}
plot_embed_chr_by_coverage <- function(all_models_unified,
                                      true_by_chr,
                                      chr_id = 1,
                                      coverage = NULL,          # NULL = all coverages
                                      embed_method = c("umap", "tsne"),
                                      normalize = TRUE,
                                      norm_args = list(per_cell_scale=TRUE, log1p_transform=TRUE, zscore_features=TRUE),
                                      n_neighbors = 10,
                                      tsne_perplexity = NULL,
                                      nstart = 10,
                                      seed = 12345,
                                      model_levels = NULL,      # e.g., final_levels
                                      ncol = 4,
                                      point_size = 1) {

  embed_method <- match.arg(embed_method)
  cell_type_label <- get_cell_type_label(chr_id, true_by_chr)

  chr_dat <- all_models_unified %>% filter(chr == chr_id)

  if (!is.null(coverage)) {
    chr_dat <- chr_dat %>% filter(coverage == coverage)
    if (nrow(chr_dat) == 0) stop("No rows for chr=", chr_id, " and coverage=", coverage)
  }

  # Determine coverage levels to iterate
  cov_levels <- levels(chr_dat$coverage)
  if (is.null(cov_levels)) cov_levels <- sort(unique(chr_dat$coverage))
  if (!is.null(coverage)) cov_levels <- unique(chr_dat$coverage)

  for (covg in cov_levels) {

    dat_cov <- chr_dat %>%
      filter(coverage == covg)

    if (!is.null(model_levels)) {
      dat_cov <- dat_cov %>%
        mutate(model = factor(model, levels = model_levels)) %>%
        arrange(model)
    } else {
      dat_cov <- dat_cov %>% arrange(model)
    }

    list_muS_cov <- setNames(dat_cov$muS, as.character(dat_cov$model))

    plots <- lapply(seq_along(list_muS_cov), function(i) {
      mat <- list_muS_cov[[i]]
      model_name <- names(list_muS_cov)[i]

      # case 1: matrix itself constant/all zero
      if (all(mat == 0) || stats::sd(as.vector(mat)) == 0) {
        return(
          ggplot() +
            geom_rect(aes(xmin = -10, xmax = 10, ymin = -10, ymax = 10),
                      fill = NA, color = "grey70", linewidth = 0.6) +
            annotate("text", x = 0, y = 0,
                     label = "(constant / all zero;\nembedding & ARI not computed)",
                     size = 4) +
            labs(title = model_name, subtitle = "ARI: NA") +
            xlab("X1") + ylab("X2") +
            theme_minimal() +
            theme(legend.position = "none")
        )
      }

      # optional normalization
      mat_use <- mat
      if (normalize) {
        mat_use <- do.call(normalize_unified, c(list(mat = mat_use), norm_args))
      }

      # ---- CHANGE: embed_cells may return NULL if all features become constant ----
      emb <- embed_cells(
        mat_use,
        embed_method = embed_method,
        n_neighbors = n_neighbors,
        tsne_perplexity = tsne_perplexity,
        seed = seed
      )

      # ---- CHANGE: handle NULL embedding (e.g., after dropping constant features) ----
      if (is.null(emb)) {
        return(
          ggplot() +
            geom_rect(aes(xmin = -10, xmax = 10, ymin = -10, ymax = 10),
                      fill = NA, color = "grey70", linewidth = 0.6) +
            annotate("text", x = 0, y = 0,
                     label = "(all features constant after preprocessing;\nembedding & ARI not computed)",
                     size = 4) +
            labs(title = model_name, subtitle = "ARI: NA") +
            xlab("X1") + ylab("X2") +
            theme_minimal() +
            theme(legend.position = "none")
        )
      }

      # cluster + ARI
      k <- length(unique(cell_type_label))
      set.seed(seed)
      km <- kmeans(emb, centers = k, nstart = nstart)

      ari_val <- adjustedRandIndex(km$cluster, as.integer(factor(cell_type_label)))

      emb_df <- tibble(
        X1 = emb$X1,
        X2 = emb$X2,
        Cluster = factor(km$cluster),
        Type = cell_type_label
      )

      ggplot(emb_df, aes(X1, X2, color = Type, shape = Cluster)) +
        geom_point(size = point_size) +
        labs(
          title = model_name,
          subtitle = paste0("ARI: ", round(ari_val, 3))
        ) +
        theme_minimal() +
        theme(legend.position = "none")
    })

    grid.arrange(
      grobs = plots,
      ncol = ncol,
      top = paste0(
        "chr", chr_id,
        ", coverage λ = ", covg,
        " | ", toupper(embed_method),
        " | ", ifelse(normalize, "normalized", "raw")
      )
    )
  }
}

```


```{r}
# ----------------------------------------------------------------------
# 5) Compute ARI for ALL rows in all_models_unified × (UMAP/tSNE × norm/no)
# ----------------------------------------------------------------------
calc_ari_all <- function(all_models_unified,
                         true_by_chr,
                         embed_methods = c("umap", "tsne"),
                         normalize_opts = c(FALSE, TRUE),
                         norm_args = list(per_cell_scale=TRUE, log1p_transform=TRUE, zscore_features=TRUE),
                         n_neighbors = 10,
                         tsne_perplexity = NULL,
                         nstart = 10,
                         seed = 12345,
                         model_levels = NULL,
                         region_info = NULL) {

  combos <- tidyr::crossing(
    embed_method = embed_methods,
    normalize = normalize_opts
  ) %>%
    mutate(combo = paste0(toupper(embed_method), " | ", ifelse(normalize, "norm", "raw")))

  out <- all_models_unified %>%
    mutate(row_id = row_number()) %>%
    tidyr::crossing(combos) %>%
    mutate(
      # CHANGE: compute true labels INSIDE pmap; no list-column map(chr, ...)
      ARI = pmap_dbl(
        list(muS, chr, embed_method, normalize),
        function(muS, chr, embed_method, normalize) {
          true_labels <- get_cell_type_label(chr, true_by_chr)

          compute_ari_one(
            mat = muS,
            true_labels = true_labels,
            embed_method = embed_method,
            normalize = normalize,
            norm_args = norm_args,
            n_neighbors = n_neighbors,
            tsne_perplexity = tsne_perplexity,
            nstart = nstart,
            seed = seed
          )
        }
      )
    ) %>%
    select(chr, coverage, model, row_id, embed_method, normalize, combo, ARI)

  if (!is.null(model_levels)) {
    out <- out %>% mutate(model = factor(model, levels = model_levels))
  }

  if (!is.null(region_info)) {
    out <- out %>% left_join(region_info, by = "chr")
  }

  out
}
```

```{r}
# --------------------------------------------
# 6) Boxplot comparison (4 combos), facetable
# --------------------------------------------
plot_ari_boxplot <- function(ari_long,
                             facet_by = c("model", "chr", "coverage", "none"),
                             ncol = 4) {
  facet_by <- match.arg(facet_by)

  p <- ggplot(ari_long, aes(x = combo, y = ARI)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, height = 0, alpha = 0.35, size = 0.8) +
    labs(
      x = NULL,
      y = "ARI (kmeans on 2D embedding)",
      title = "ARI comparison: UMAP vs tSNE × normalization vs none"
    ) +
    theme_classic(base_size = 12) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

  if (facet_by != "none") {
    p <- p + facet_wrap(as.formula(paste0("~", facet_by)), ncol = ncol)
  }

  p
}

```

```{r, fig.width=12}
# ============================================================
# USAGE EXAMPLES
# ============================================================

# (A) QUICK TEST: one chr, all coverages, UMAP raw
plot_embed_chr_by_coverage(all_models_unified, true_by_chr,
                           chr_id = 1,
                           embed_method = "umap",
                           normalize = FALSE,
                           n_neighbors = n_rep,
                           nstart = n_rep,
                           model_levels = final_levels)
```

```{r, fig.width=12}

# (B) QUICK TEST: one chr, one coverage level (e.g. "1e-03"), tSNE normalized
plot_embed_chr_by_coverage(all_models_unified, true_by_chr,
                           chr_id = 10,
                           coverage = 0.1,
                           embed_method = "tsne",
                           normalize = TRUE,
                           n_neighbors = n_rep,
                           nstart = n_rep,
                           model_levels = final_levels)

```


```{r}
# (C) Compute ARI for everything (10 chr × 10 cov × models × 4 combos)
ari_long <- calc_ari_all(all_models_unified, true_by_chr,
                         embed_methods = c("umap","tsne"),
                         normalize_opts = c(FALSE, TRUE),
                         norm_args = list(per_cell_scale=TRUE, log1p_transform=TRUE, zscore_features=TRUE),
                         n_neighbors = n_rep,
                         nstart = n_rep,
                         seed = 12345,
                         model_levels = final_levels,
                         region_info = region_info %>% select(chr, facet_label))

```

```{r, fig.width=15}
ari_df2 <- ari_long %>%
  mutate(
    embed_label = toupper(embed_method),
    norm_label = ifelse(normalize, "Normalized", "Unnormalized"),
    facet_label = paste(embed_label, norm_label, sep = " | ")
  )
p_ari3 <- ggplot(ari_df2,
                 aes(x = coverage,
                     y = ARI,
                     fill = model,
                     color = model)) +

  geom_boxplot(outlier.size = 1,
               width = 0.3,
               position = position_dodge(width = 0.6),
               linewidth = 0.3) +

  facet_grid( embed_label ~ norm_label) +

  labs(
    x = expression(lambda),
    y = "ARI",
    title = "ARI Values in Clustering Across Coverage Levels",
    subtitle = "Comparison of UMAP vs tSNE with and without normalization"
  ) +

  theme_classic(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey95", color = NA)
  )

p_ari3

```

## Try more normalization options:

```{r}
# 0) helper to run one setting and label it
run_one_norm <- function(norm_label,
                         normalize_opts,
                         norm_args = NULL) {

  calc_ari_all(
    all_models_unified, true_by_chr,
    embed_methods = c("umap", "tsne"),
    normalize_opts = normalize_opts,
    norm_args = if (is.null(norm_args)) list(per_cell_scale=TRUE, log1p_transform=TRUE, zscore_features=TRUE) else norm_args,
    n_neighbors = n_rep,
    nstart = n_rep,
    seed = 12345,
    model_levels = final_levels,
    region_info = region_info %>% select(chr, facet_label)
  ) %>%
    mutate(norm_label = norm_label)
}

# A) Raw (no normalization)
ari_raw <- run_one_norm(
  norm_label = "None",
  normalize_opts = c(FALSE)
)

# B) Log1p only
ari_log1p <- run_one_norm(
  norm_label = "Log1p",
  normalize_opts = c(TRUE),
  norm_args = list(per_cell_scale = FALSE, log1p_transform = TRUE, zscore_features = FALSE)
)

# C) Cell-scale + Log1p
ari_scale_log1p <- run_one_norm(
  norm_label = "CellScale+Log1p",
  normalize_opts = c(TRUE),
  norm_args = list(per_cell_scale = TRUE, log1p_transform = TRUE, zscore_features = FALSE)
)

# D) Full (your current "normalized")
ari_full <- run_one_norm(
  norm_label = "Full (Scale+Log1p+Z)",
  normalize_opts = c(TRUE),
  norm_args = list(per_cell_scale = TRUE, log1p_transform = TRUE, zscore_features = TRUE)
)

# Combine
ari_all4 <- bind_rows(ari_raw, ari_log1p, ari_scale_log1p, ari_full) %>%
  mutate(embed_label = toupper(embed_method))
```

```{r, fig.width=15}
ari_all4$norm_label = factor(ari_all4$norm_label, levels = c("None", "Log1p", "CellScale+Log1p", "Full (Scale+Log1p+Z)"))
ggplot(ari_all4,
                 aes(x = coverage,
                     y = ARI,
                     fill = model,
                     color = model)) +

  geom_boxplot(outlier.size = 1,
               width = 0.3,
               position = position_dodge(width = 0.6),
               linewidth = 0.3) +
# stat_summary(fun = median,
#              geom = "point",
#              position = position_dodge(width = 0.8),
#              size = 1.8,
#              shape = 95) +  # thick horizontal line
  facet_grid( embed_label ~ norm_label) +

  labs(
    x = expression(lambda),
    y = "ARI",
    title = "ARI Values in Clustering Across Coverage Levels",
    subtitle = "Comparison of UMAP vs tSNE with and without normalization"
  ) +

  theme_classic(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey95", color = NA)
  )

```

## Try stratum and truncation

```{r}
n_bins = 100
# Compute genomic distance (in bins) for each (i,j) pair
# distance (in bins) for each upper-tri pair row (N x K matrix)
compute_stratum_index <- function(n_bins) {
  ij <- which(upper.tri(matrix(0, n_bins, n_bins)), arr.ind = TRUE)
  abs(ij[,2] - ij[,1])
}

# 50kb bins, truncate to <2Mb => 2,000,000 / 50,000 = 40 bins
max_bin_dist <- 40
```


```{r}
bandnorm_zscore <- function(mat, stratum_index) {
  mat2 <- mat
  for (d in sort(unique(stratum_index))) {
    idx <- which(stratum_index == d)
    if (length(idx) < 2) next
    sub <- mat2[idx, , drop = FALSE]
    subz <- t(scale(t(sub)))
    subz[is.na(subz)] <- 0
    mat2[idx, ] <- subz
  }
  mat2
}

oe_by_band <- function(mat, stratum_index, eps = 1e-12) {
  mat2 <- mat
  for (d in sort(unique(stratum_index))) {
    idx <- which(stratum_index == d)
    if (length(idx) < 2) next
    exp_d <- mean(mat2[idx, ], na.rm = TRUE)
    if (is.finite(exp_d) && exp_d > eps) {
      mat2[idx, ] <- mat2[idx, ] / exp_d
    }
  }
  mat2
}

truncate_by_distance <- function(mat, stratum_index, max_bin_dist = 40) {
  keep <- which(stratum_index <= max_bin_dist)
  mat[keep, , drop = FALSE]
}
```

```{r}
# Build a distance index ONCE (based on N of your matrices)
n_bins <- 100
stratum_index_full <- compute_stratum_index(n_bins)

# For truncation we also need truncated stratum index
keep_trunc <- which(stratum_index_full <= max_bin_dist)
stratum_index_trunc <- stratum_index_full[keep_trunc]

# Define transform functions (input: mat N x K, output: transformed mat)
dist_norm_funs <- list(
  "BandNorm" = function(mat) {
    bandnorm_zscore(mat, stratum_index_full)
  },
  "OE" = function(mat) {
    oe_by_band(mat, stratum_index_full)
  },
  "Trunc(<2Mb)" = function(mat) {
    truncate_by_distance(mat, stratum_index_full, max_bin_dist)
  },
  "Trunc+BandNorm" = function(mat) {
    mat2 <- truncate_by_distance(mat, stratum_index_full, max_bin_dist)
    bandnorm_zscore(mat2, stratum_index_trunc)
  },
  "Trunc+OE" = function(mat) {
    mat2 <- truncate_by_distance(mat, stratum_index_full, max_bin_dist)
    oe_by_band(mat2, stratum_index_trunc)
  }
)
```


```{r}
calc_ari_all_transform <- function(all_models_unified,
                                   true_by_chr,
                                   embed_methods = c("umap","tsne"),
                                   transform_name = "BandNorm",
                                   transform_fun,
                                   n_neighbors = 10,
                                   tsne_perplexity = NULL,
                                   nstart = 10,
                                   seed = 12345,
                                   model_levels = NULL,
                                   region_info = NULL) {

  combos <- tidyr::crossing(embed_method = embed_methods) %>%
    mutate(embed_label = toupper(embed_method),
           norm_label = transform_name)

  out <- all_models_unified %>%
    mutate(row_id = row_number()) %>%
    tidyr::crossing(combos) %>%
    mutate(
      ARI = pmap_dbl(
        list(muS, chr, embed_method),
        function(muS, chr, embed_method) {

          true_labels <- get_cell_type_label(chr, true_by_chr)

          # apply distance-aware transform
          mat_use <- transform_fun(muS)

          # embed (your embed_cells already handles constant-feature issues)
          emb <- embed_cells(
            mat_use,
            embed_method = embed_method,
            n_neighbors = n_neighbors,
            tsne_perplexity = tsne_perplexity,
            seed = seed
          )
          if (is.null(emb)) return(NA_real_)

          k <- length(unique(true_labels))
          set.seed(seed)
          km <- kmeans(emb, centers = k, nstart = nstart)

          adjustedRandIndex(km$cluster, as.integer(factor(true_labels)))
        }
      )
    ) %>%
    transmute(chr, coverage, model, row_id,
              embed_method,
              embed_label,
              norm_label,
              ARI)

  if (!is.null(model_levels)) out <- out %>% mutate(model = factor(model, levels = model_levels))
  if (!is.null(region_info))  out <- out %>% left_join(region_info, by = "chr")
  out
}
```

```{r}
ari_dist <- purrr::imap_dfr(dist_norm_funs, function(fun, nm) {
  calc_ari_all_transform(
    all_models_unified, true_by_chr,
    embed_methods = c("umap","tsne"),
    transform_name = nm,
    transform_fun = fun,
    n_neighbors = n_rep,
    nstart = n_rep,
    seed = 12345,
    model_levels = final_levels,
    region_info = region_info %>% select(chr, facet_label)
  )
})

# Combine global + distance-aware
ari_compare_all <- bind_rows(
  ari_all4 %>% mutate(embed_label = toupper(embed_method)),
  ari_dist
)
```

```{r}
ari_compare_all$norm_label %>% table()
```


```{r, fig.width=15}
p_compare <- ggplot(ari_compare_all %>% filter(norm_label %in% c("None", "Log1p", "BandNorm", "OE")) %>%
                      mutate(norm_label = factor(norm_label, levels = c("None", "Log1p", "BandNorm", "OE"))),
                    aes(x = coverage, y = ARI, fill = model, color = model)) +
  geom_boxplot(outlier.size = 0.8,
               width = 0.6,
               position = position_dodge(width = 0.8),
               linewidth = 0.6,
               alpha = 0.7) +
  facet_grid( embed_label ~ norm_label) +
  labs(
    x = expression(lambda),
    y = "ARI",
    title = "ARI across coverage levels",
    subtitle = "Global vs distance-aware normalizations; UMAP vs tSNE"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey95", color = NA)
  )

p_compare
```

```{r, fig.width=15}
p_compare <- ggplot(ari_compare_all %>% filter(norm_label %in% c("None", "Log1p", "Trunc(<2Mb)", "OE", "Trunc+OE")) %>%
                      mutate(norm_label = factor(norm_label, levels = c("None", "Log1p", "Trunc(<2Mb)", "OE", "Trunc+OE"))),
                    aes(x = coverage, y = ARI, fill = model, color = model)) +
  geom_boxplot(outlier.size = 0.8,
               width = 0.6,
               position = position_dodge(width = 0.8),
               linewidth = 0.6,
               alpha = 0.7) +
  facet_grid( embed_label ~ norm_label) +
  labs(
    x = expression(lambda),
    y = "ARI",
    title = "ARI across coverage levels",
    subtitle = "Global vs distance-aware normalizations; UMAP vs tSNE"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey95", color = NA)
  )

p_compare
```

```{r, fig.width=15}
ggplot(ari_compare_all,
                    aes(x = norm_label, y = ARI, fill = model, color = model)) +
  geom_boxplot(outlier.size = 0.8,
               width = 0.6,
               position = position_dodge(width = 0.8),
               linewidth = 0.6,
               alpha = 0.7) +
  facet_grid(~embed_label) +
  labs(
    x = expression(lambda),
    y = "ARI",
    title = "ARI across coverage levels",
    subtitle = "Global vs distance-aware normalizations; UMAP vs tSNE"
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey95", color = NA)
  )

```



#### Embedding Methods

- UMAP and tSNE showed broadly comparable performance across all simulation settings.
- No consistent or substantial advantage of one embedding method over the other was observed after aggregating results across chromosomes and coverage levels.
- Therefore, embedding choice was not the primary driver of clustering performance differences.

#### Normalization Methods Evaluated 


- None (raw)
- Log1p transformation
- CellScale + Log1p
- Full normalization (Scale + Log1p + Z-score)

- BandNorm (distance-stratum z-score)
- Observed/Expected (O/E)
- Truncated at <2Mb (40bins), only keeping near diagonal elements

#### Key Findings

- Log1p and O/E were the only normalization methods that consistently improved ARI compared to no normalization.
- Full normalization (including z-scoring) often reduced performance, suggesting that over-normalization removed biologically meaningful signal.
- BandNorm and truncation-based strategies did not provide consistent improvements when performance was aggregated across all coverage levels.
- Log1p performed as well as or slightly better than O/E while being simpler and more stable across models.


