---
title: "Server InnerProduct: SCORE JSON vs Recomputed (Embedding32/PCA2D)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1) Packages

```{r packages}
required_pkgs <- c(
  "dplyr", "tidyr", "ggplot2", "readr", "purrr", "tibble", "jsonlite",
  "hdf5r", "Matrix", "cluster", "mclust", "aricode"
)
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop("Missing R packages: ", paste(missing_pkgs, collapse = ", "))
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(purrr)
library(tibble)
library(jsonlite)
library(hdf5r)
library(Matrix)
library(cluster)
library(mclust)
library(aricode)
```

## 2) Inputs (Server InnerProduct Only)

```{r inputs}
n_runs <- 10L
seed_start <- 2026L

h5ad_root <- "D:/Codex/Test/downloaded_results/server_json_runs"

run_tbl <- tidyr::expand_grid(
  method = c("raw", "bhzip"),
  run = seq_len(n_runs)
) %>%
  mutate(
    seed = seed_start + run - 1L,
    dset = sprintf("oocyte_zygote_%s_pairs_r%02d", method, run),
    json_path = file.path(h5ad_root, dset, "1M", "innerproduct", "innerproduct.json"),
    json_exists = file.exists(json_path),
    h5ad = file.path(h5ad_root, dset, "1M", "innerproduct", "anndata_obj.h5ad"),
    h5ad_exists = file.exists(h5ad)
  )

if (any(!run_tbl$json_exists)) {
  print(run_tbl %>% filter(!json_exists))
  stop("Missing one or more innerproduct.json files under: ", h5ad_root)
}

if (any(!run_tbl$h5ad_exists)) {
  print(run_tbl %>% filter(!h5ad_exists))
  stop("Missing one or more anndata_obj.h5ad files under: ", h5ad_root)
}
```

## 3) Helpers

```{r helpers}
read_h5ad_csr <- function(g) {
  data <- as.numeric(g[["data"]]$read())
  indices <- as.integer(g[["indices"]]$read()) + 1L
  indptr <- as.integer(g[["indptr"]]$read())
  shape <- as.integer(g[["shape"]]$read())
  n_row <- shape[1]
  n_col <- shape[2]
  i <- rep(seq_len(n_row), diff(indptr))
  Matrix::sparseMatrix(i = i, j = indices, x = data, dims = c(n_row, n_col))
}

read_h5ad_matrix <- function(node) {
  if (inherits(node, "H5D")) {
    return(as.matrix(node$read()))
  }
  if (inherits(node, "H5Group")) {
    nms <- names(node)
    if (all(c("data", "indices", "indptr", "shape") %in% nms)) {
      return(as.matrix(read_h5ad_csr(node)))
    }
  }
  stop("Unsupported matrix encoding in h5ad.")
}

read_h5ad_obs_col <- function(obs_group, key) {
  node <- obs_group[[key]]
  if (inherits(node, "H5Group")) {
    nms <- names(node)
    if (all(c("codes", "categories") %in% nms)) {
      codes <- as.integer(node[["codes"]]$read())
      categories <- as.character(node[["categories"]]$read())
      return(as.vector(categories[codes + 1L]))
    }
  }
  if (inherits(node, "H5D")) {
    vals <- node$read()
    attrs <- tryCatch(hdf5r::h5attr_names(node), error = function(e) character(0))
    if ("categories" %in% attrs) {
      cats_obj <- hdf5r::h5attr(node, "categories")
      categories <- as.character(cats_obj$dereference()[[1]]$read())
      return(as.vector(categories[as.integer(vals) + 1L]))
    }
    return(as.vector(as.character(vals)))
  }
  stop("Unsupported obs encoding for key: ", key)
}

read_h5ad_run <- function(path) {
  h5 <- hdf5r::H5File$new(path, mode = "r")
  on.exit(try(h5$close_all(), silent = TRUE), add = TRUE)

  X <- read_h5ad_matrix(h5[["X"]])
  X <- as.matrix(X)
  storage.mode(X) <- "double"

  X_pca <- NULL
  if ("obsm" %in% names(h5) && "X_pca" %in% names(h5[["obsm"]])) {
    X_pca <- read_h5ad_matrix(h5[["obsm"]][["X_pca"]])
    X_pca <- as.matrix(X_pca)
    storage.mode(X_pca) <- "double"
  }

  y <- factor(as.vector(read_h5ad_obs_col(h5[["obs"]], "celltype")))
  list(X = X, X_pca = X_pca, y = y)
}

orient_rows_cells <- function(mat, y, label) {
  if (is.null(mat)) return(NULL)
  if (nrow(mat) == length(y)) return(mat)
  if (ncol(mat) == length(y)) return(t(mat))
  stop(sprintf(
    "%s shape mismatch: matrix is %d x %d, but labels length is %d",
    label, nrow(mat), ncol(mat), length(y)
  ))
}

cosine_dist <- function(z) {
  z <- as.matrix(z)
  nrm <- sqrt(rowSums(z^2))
  nrm[nrm == 0] <- 1
  zn <- z / nrm
  sim <- tcrossprod(zn)
  sim[sim > 1] <- 1
  sim[sim < -1] <- -1
  as.dist(1 - sim)
}

silhouette_cosine <- function(z, labels) {
  labs <- as.integer(factor(labels))
  if (length(unique(labs)) < 2) return(NA_real_)
  s <- cluster::silhouette(labs, dist = cosine_dist(z))
  mean(s[, "sil_width"], na.rm = TRUE)
}

ensure_min_2d <- function(z) {
  z <- as.matrix(z)
  if (ncol(z) >= 2) return(z)
  cbind(z, rep(0, nrow(z)))
}

safe_gmm_labels <- function(z, n_clusters, seed) {
  z2 <- ensure_min_2d(z)
  set.seed(seed)
  fit <- tryCatch(
    mclust::Mclust(z2, G = n_clusters, verbose = FALSE),
    error = function(e) NULL
  )
  if (!is.null(fit) && !is.null(fit$classification)) {
    return(as.integer(fit$classification))
  }
  as.integer(kmeans(z2, centers = n_clusters, nstart = 10, iter.max = 100)$cluster)
}

compute_space_metrics <- function(z, y, seed) {
  set.seed(seed)
  z <- as.matrix(z)
  storage.mode(z) <- "double"
  y <- factor(as.vector(y))
  if (length(y) != nrow(z)) stop("Label length does not match number of rows in matrix.")
  n_clusters <- length(unique(y))

  pred_kmeans <- kmeans(z, centers = n_clusters, nstart = 10, iter.max = 100)$cluster
  pred_agglom <- cutree(hclust(dist(z), method = "ward.D2"), k = n_clusters)
  pred_gmm <- safe_gmm_labels(z, n_clusters, seed)

  out <- list(
    "ari_k-means" = mclust::adjustedRandIndex(y, pred_kmeans),
    "ari_agglomerative" = mclust::adjustedRandIndex(y, pred_agglom),
    "ari_gmm" = mclust::adjustedRandIndex(y, pred_gmm),
    "nmi_k-means" = aricode::NMI(as.integer(y), as.integer(pred_kmeans)),
    "nmi_agglomerative" = aricode::NMI(as.integer(y), as.integer(pred_agglom)),
    "nmi_gmm" = aricode::NMI(as.integer(y), as.integer(pred_gmm)),
    "silhouette_k-means" = silhouette_cosine(z, pred_kmeans),
    "silhouette_agglomerative" = silhouette_cosine(z, pred_agglom),
    "silhouette_gmm" = silhouette_cosine(z, pred_gmm)
  )

  out$best_ari <- max(unlist(out[c("ari_k-means", "ari_agglomerative", "ari_gmm")]), na.rm = TRUE)
  out$best_nmi <- max(unlist(out[c("nmi_k-means", "nmi_agglomerative", "nmi_gmm")]), na.rm = TRUE)
  out$best_silhouette <- max(unlist(out[c("silhouette_k-means", "silhouette_agglomerative", "silhouette_gmm")]), na.rm = TRUE)
  out$avgBio <- mean(c(out$best_ari, out$best_nmi, out$best_silhouette), na.rm = TRUE)

  tibble::as_tibble(out)
}

first_numeric <- function(x) {
  if (is.null(x)) return(NA_real_)
  if (is.list(x)) x <- unlist(x, recursive = TRUE, use.names = FALSE)
  vals <- suppressWarnings(as.numeric(x))
  if (length(vals) == 0 || all(is.na(vals))) return(NA_real_)
  vals[which(!is.na(vals))[1]]
}

extract_relevant_metrics_from_json <- function(path) {
  if (is.na(path) || !file.exists(path)) {
    return(tibble(metric = character(), value = numeric()))
  }

  j <- jsonlite::fromJSON(path, simplifyVector = FALSE)
  keys <- names(j)
  if (is.null(keys) || length(keys) == 0) {
    return(tibble(metric = character(), value = numeric()))
  }

  vals <- purrr::map_dbl(keys, ~ first_numeric(j[[.x]]))
  base_df <- tibble(metric = keys, value = vals) %>%
    filter(!is.na(value)) %>%
    filter(grepl("(ari|nmi|silhouette)", metric, ignore.case = TRUE)) %>%
    filter(!grepl("silhouette-gt", metric, ignore.case = TRUE))

  best_ari <- first_numeric(j[["best_ari"]])
  best_nmi <- first_numeric(j[["best_nmi"]])
  best_asw <- first_numeric(j[["best_silhouette"]])
  avgBio <- mean(c(best_ari, best_nmi, best_asw), na.rm = TRUE)
  if (all(is.na(c(best_ari, best_nmi, best_asw)))) avgBio <- NA_real_

  bind_rows(
    base_df,
    tibble(metric = "avgBio", value = avgBio)
  )
}
```

## 4) Build Plot Data from 3 Sources

```{r data_build}
# Source A: SCORE JSON (same style as extract_paper_metrics...)
id_cols <- c("source_option", "method", "run", "seed")
score_json_long <- run_tbl %>%
  transmute(method, run, seed, json_path) %>%
  mutate(metrics = purrr::map(json_path, extract_relevant_metrics_from_json)) %>%
  tidyr::unnest(metrics) %>%
  mutate(source_option = "SCORE_JSON") %>%
  select(all_of(id_cols), metric, value)

# Source B/C: Recompute from h5ad
recomp_df <- purrr::pmap_dfr(
  list(run_tbl$method, run_tbl$run, run_tbl$seed, run_tbl$h5ad),
  function(method, run, seed, h5ad) {
    dat <- read_h5ad_run(h5ad)
    y <- dat$y
    x <- orient_rows_cells(dat$X, y, "X")

    z_pca <- if (!is.null(dat$X_pca)) {
      x_pca <- orient_rows_cells(dat$X_pca, y, "X_pca")
      x_pca[, seq_len(min(2L, ncol(x_pca))), drop = FALSE]
    } else {
      stats::prcomp(x, center = TRUE, scale. = FALSE)$x[, 1:2, drop = FALSE]
    }
    z_pca <- ensure_min_2d(z_pca)

    m_emb <- compute_space_metrics(x, y, seed) %>%
      pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
      mutate(source_option = "Recomputed_Embedding32")

    m_pca <- compute_space_metrics(z_pca, y, seed) %>%
      pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
      mutate(source_option = "Recomputed_PCA2D")

    bind_rows(m_emb, m_pca) %>%
      mutate(method = method, run = run, seed = seed) %>%
      select(all_of(id_cols), metric, value)
  }
)

plot_df0 <- bind_rows(score_json_long, recomp_df)

# Keep paper metric families; exclude silhouette-gt
plot_df0 <- plot_df0 %>%
  filter(grepl("(ari|nmi|silhouette|avgBio)", metric, ignore.case = TRUE)) %>%
  filter(!grepl("silhouette-gt", metric, ignore.case = TRUE))

# Keep only metrics present in all sources (fair side-by-side comparison)
common_metrics <- plot_df0 %>%
  group_by(source_option, metric) %>%
  summarise(n = sum(!is.na(value)), .groups = "drop") %>%
  group_by(metric) %>%
  summarise(n_sources = sum(n > 0), .groups = "drop") %>%
  filter(n_sources == 3) %>%
  pull(metric)

metric_order <- c(
  "ari_agglomerative", "ari_gmm", "ari_k-means", "best_ari",
  "nmi_agglomerative", "nmi_gmm", "nmi_k-means", "best_nmi",
  "silhouette_agglomerative", "silhouette_gmm", "silhouette_k-means", "best_silhouette",
  "avgBio"
)

metric_order <- metric_order[metric_order %in% common_metrics]

plot_df <- plot_df0 %>%
  filter(metric %in% metric_order) %>%
  mutate(
    metric = factor(metric, levels = metric_order),
    method = factor(method, levels = c("raw", "bhzip"), labels = c("Raw", "HiCBZIP-GB(NB)")),
    source_option = factor(
      source_option,
      levels = c("SCORE_JSON", "Recomputed_Embedding32", "Recomputed_PCA2D")
    )
  ) %>%
  arrange(source_option, method, run, metric)

source_labeller <- c(
  "SCORE_JSON" = "Original SCORE 32D",
  "Recomputed_Embedding32" = "Recomputed 32D",
  "Recomputed_PCA2D" = "Recomputed PCA2D"
)

coverage_tbl <- plot_df %>%
  group_by(source_option) %>%
  summarise(found = n_distinct(paste(method, run, seed)), .groups = "drop")
print(coverage_tbl)
```

## 5) Summary Table

```{r summary}
summary_df <- plot_df %>%
  group_by(source_option, method, metric) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    .groups = "drop"
  ) %>%
  arrange(source_option, metric, method)

summary_fmt <- summary_df %>%
  mutate(mean_sd = sprintf("%.3f (%.3f)", mean, sd)) %>%
  select(source_option, method, metric, mean_sd, n)

print(summary_fmt, n = 100)
```

## 6) Plot (Facet Rows by Source Option)

```{r plot, fig.width=15, fig.height=10}
group_boundaries <- tibble::tibble(
  metric = levels(plot_df$metric),
  idx = seq_along(levels(plot_df$metric)),
  metric_group = dplyr::case_when(
    grepl("^ari|best_ari$", metric, ignore.case = TRUE) ~ "ARI",
    grepl("^nmi|best_nmi$", metric, ignore.case = TRUE) ~ "NMI",
    grepl("^silhouette|best_silhouette$", metric, ignore.case = TRUE) ~ "Silhouette",
    metric == "avgBio" ~ "AvgBIO",
    TRUE ~ "Other"
  )
)

separator_x <- group_boundaries %>%
  group_by(metric_group) %>%
  summarise(last_idx = max(idx), .groups = "drop") %>%
  arrange(last_idx)

if (nrow(separator_x) > 1) {
  separator_x <- separator_x$last_idx[seq_len(nrow(separator_x) - 1)] + 0.5
} else {
  separator_x <- numeric(0)
}

label_x <- group_boundaries %>%
  group_by(metric_group) %>%
  summarise(x = mean(idx), .groups = "drop")

label_df <- tidyr::crossing(
  source_option = levels(plot_df$source_option),
  label_x
) %>%
  mutate(y_group_label = 1)

p_rows <- ggplot(plot_df, aes(x = metric, y = value, fill = method, color = method)) +
  geom_boxplot(
    width = 0.68,
    alpha = 0.30,
    linewidth = 0.8,
    outlier.shape = NA,
    position = position_dodge(width = 0.78)
  ) +
  geom_point(
    alpha = 0.60,
    size = 1.2,
    position = position_jitterdodge(dodge.width = 0.78, jitter.width = 0.10)
  ) +
  geom_vline(
    xintercept = separator_x,
    linetype = "dashed",
    color = "grey45",
    linewidth = 0.5
  ) +
  geom_text(
    data = label_df,
    aes(x = x, y = y_group_label, label = metric_group),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.2
  ) +
  facet_grid(source_option ~ ., scales = "fixed", labeller = as_labeller(source_labeller)) +
  scale_fill_manual(values = c("Raw" = "#4E79A7", "HiCBZIP-GB(NB)" = "#E15759")) +
  scale_color_manual(values = c("Raw" = "#2F5D8A", "HiCBZIP-GB(NB)" = "#B63D40")) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  labs(
    title = "Server InnerProduct: Source Comparison",
    subtitle = "Original SCORE 32D vs Recomputed 32D vs Recomputed PCA2D",
    x = "Metric",
    y = "Value",
    fill = "Input",
    color = "Input"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey95"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "top"
  )

print(p_rows)
```

## 7) Metric-by-Source Plot (Compare Sources + Model Input)

```{r plot_source_metric_model, fig.width=16, fig.height=10}
p_source_metric <- ggplot(
  plot_df,
  aes(x = source_option, y = value, fill = method, color = method)
) +
  geom_boxplot(
    width = 0.68,
    alpha = 0.30,
    linewidth = 0.8,
    outlier.shape = NA,
    position = position_dodge(width = 0.78)
  ) +
  geom_point(
    alpha = 0.60,
    size = 1.2,
    position = position_jitterdodge(dodge.width = 0.78, jitter.width = 0.10)
  ) +
  facet_wrap(~ metric, ncol = 4, scales = "fixed") +
  scale_x_discrete(labels = source_labeller) +
  scale_fill_manual(values = c("Raw" = "#4E79A7", "HiCBZIP-GB(NB)" = "#E15759")) +
  scale_color_manual(values = c("Raw" = "#2F5D8A", "HiCBZIP-GB(NB)" = "#B63D40")) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Metric Comparison by Evaluation Source",
    subtitle = "Evaluation sources: Original SCORE 32D, Recomputed 32D, Recomputed PCA2D; model inputs: Raw vs HiCBZIP-GB(NB)",
    x = "Evaluation Source",
    y = "Metric Value",
    fill = "Input",
    color = "Input"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey95"),
    axis.text.x = element_text(angle = 20, hjust = 1),
    legend.position = "top"
  )

print(p_source_metric)
```


- Recomputed 32D uses the same implementation R code for clustering as PCA2D 

- PCA2D shows much higher silhouette but similar ARI/NMI. But lower dimensional should naturally have higher silhouette?

- So basically PCA2D did not improve clustering but will loose some information in 32D.