---
title: "Paper Metrics: External Imputations + Integrated Methods with Fast-Higashi and scHiCluster Heatmap"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1) Packages

```{r packages}
required_pkgs <- c("dplyr", "tidyr", "ggplot2", "readr", "stringr", "purrr", "tibble", "hdf5r", "patchwork", "reticulate")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop("Missing R packages: ", paste(missing_pkgs, collapse = ", "))
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(stringr)
library(purrr)
library(tibble)
library(hdf5r)
library(patchwork)
```

## 2) Inputs

```{r inputs}
n_runs <- 10L
seed_start <- 2026L

innerproduct_nb_root <- "D:/Codex/Test/downloaded_results/server_json_runs"
innerproduct_nm_root <- "D:/Codex/Test/downloaded_results/server_json_runs_nm"
innerproduct_schicimpute_root <- "D:/Codex/Test/downloaded_results/oocyte_zygote_schicimpute_10runs_innerproduct"

snapatac_noidf_nb_root <- "D:/Codex/Test/downloaded_results/server_json_runs"
snapatac_noidf_nm_root <- "D:/Codex/Test/downloaded_results/server_json_runs_nm"
snapatac_noidf_schicimpute_root <- "D:/Codex/Test/downloaded_results/oocyte_zygote_schicimpute_10runs_snapatac_noidf"

schicluster_root <- "D:/Codex/Test/downloaded_results/oocyte_zygote_schicluster_10runs"
fast_higashi_root <- "D:/Codex/Test/downloaded_results/oocyte_zygote_fast_higashi_10runs"
root_nb <- "D:/Codex/Test/downloaded_results/server_json_runs"
root_nm <- "D:/Codex/Test/downloaded_results/server_json_runs_nm"

metric_order <- c(
  "ari_agglomerative", "ari_gmm", "ari_k-means", "ari_leiden", "ari_louvain", "best_ari",
  "nmi_agglomerative", "nmi_gmm", "nmi_k-means", "nmi_leiden", "nmi_louvain", "best_nmi",
  "silhouette_agglomerative", "silhouette_gmm", "silhouette_k-means",
  "silhouette_leiden", "silhouette_louvain", "best_silhouette",
  "avgBio"
)

external_inputs <- c(
  raw = "Raw",
  bhzip = "HiCBZIP-GB(NB)",
  bhzip_nm = "HiCBZIP-N(M)",
  schicimpute = "scHiCImpute"
)

base_metric_cols <- c(
  "ari_k-means", "ari_agglomerative", "ari_gmm", "ari_louvain", "ari_leiden", "best_ari",
  "nmi_k-means", "nmi_agglomerative", "nmi_gmm", "nmi_louvain", "nmi_leiden", "best_nmi",
  "silhouette_k-means", "silhouette_agglomerative", "silhouette_gmm",
  "silhouette_louvain", "silhouette_leiden", "best_silhouette", "best_silhouette-gt", "wall_time"
)

pivot_metric_cols <- c(base_metric_cols[base_metric_cols != "wall_time"], "avgBio")
```

## 3) Resolve JSON Inputs

```{r read_jsons}
extract_metric_json <- function(path, key) {
  if (is.na(path) || !file.exists(path)) return(NA_real_)
  x <- jsonlite::fromJSON(path, simplifyVector = FALSE)[[key]]
  if (is.null(x)) return(NA_real_)
  if (is.list(x)) x <- unlist(x, recursive = TRUE, use.names = FALSE)
  vals <- suppressWarnings(as.numeric(x))
  if (length(vals) == 0 || all(is.na(vals))) return(NA_real_)
  vals[which(!is.na(vals))[1]]
}

resolve_json_path <- function(root_string, dset_name, alg_dir, json_name) {
  roots <- strsplit(root_string, "\\|")[[1]]
  roots <- roots[nzchar(roots)]
  alg_dir_alt <- gsub(":", "%3A", alg_dir, fixed = TRUE)
  json_name_alt <- gsub(":", "%3A", json_name, fixed = TRUE)
  candidates <- unlist(lapply(roots, function(rt) {
    c(
      file.path(rt, dset_name, "1M", alg_dir, json_name),
      file.path(rt, dset_name, "1M", alg_dir_alt, json_name),
      file.path(rt, dset_name, "1M", alg_dir_alt, json_name_alt),
      file.path(rt, dset_name, "1M", alg_dir, sprintf("%s_1M_%s", dset_name, json_name)),
      file.path(rt, dset_name, "1M", alg_dir_alt, sprintf("%s_1M_%s", dset_name, json_name_alt)),
      file.path(rt, alg_dir, dset_name, json_name)
    )
  }))
  hit <- candidates[file.exists(candidates)]
  if (length(hit) > 0) hit[[1]] else NA_character_
}

path_tbl <- bind_rows(
  tibble::tribble(
    ~run_option,       ~input_code,    ~input_label,         ~roots_chr,                                                       ~dset_template,                                      ~alg_dir,                                         ~json_name,                                                    ~family,
    "InnerProduct",    "raw",          "Raw",                innerproduct_nb_root,                                              "oocyte_zygote_raw_pairs_r%02d",                    "innerproduct",                                  "innerproduct.json",                                         "External imputation + downstream embedding",
    "InnerProduct",    "bhzip",        "HiCBZIP-GB(NB)",     innerproduct_nb_root,                                              "oocyte_zygote_bhzip_pairs_r%02d",                  "innerproduct",                                  "innerproduct.json",                                         "External imputation + downstream embedding",
    "InnerProduct",    "bhzip_nm",     "HiCBZIP-N(M)",       innerproduct_nm_root,                                              "oocyte_zygote_bhzip_nm_pairs_r%02d",               "innerproduct",                                  "innerproduct.json",                                         "External imputation + downstream embedding",
    "InnerProduct",    "schicimpute",  "scHiCImpute",        innerproduct_schicimpute_root,                                     "oocyte_zygote_schicimpute_pairs_r%02d",            "innerproduct",                                  "innerproduct.json",                                         "External imputation + downstream embedding",
    "SnapATAC_noIDF",  "raw",          "Raw",                snapatac_noidf_nm_root,                                            "oocyte_zygote_raw_nm_pairs_snapatac_no_idf_r%02d", "snapatac",                                      "snapatac.json",                                             "External imputation + downstream embedding",
    "SnapATAC_noIDF",  "bhzip",        "HiCBZIP-GB(NB)",     paste(snapatac_noidf_nb_root, snapatac_noidf_nm_root, sep = "|"), "oocyte_zygote_bhzip_pairs_snapatac_no_idf_gb_only_r%02d", "snapatac",                              "snapatac.json",                                             "External imputation + downstream embedding",
    "SnapATAC_noIDF",  "bhzip_nm",     "HiCBZIP-N(M)",       snapatac_noidf_nm_root,                                            "oocyte_zygote_bhzip_nm_pairs_snapatac_no_idf_r%02d", "snapatac",                                    "snapatac.json",                                             "External imputation + downstream embedding",
    "SnapATAC_noIDF",  "schicimpute",  "scHiCImpute",        snapatac_noidf_schicimpute_root,                                   "oocyte_zygote_schicimpute_pairs_r%02d",            "snapatac",                                      "snapatac.json",                                             "External imputation + downstream embedding",
    "scHiCluster",     "integrated",   "scHiCluster",        schicluster_root,                                                  "oocyte_zygote_schicluster_r%02d",                  "schicluster:vc_sqrt_norm,convolution,random_walk", "schicluster:vc_sqrt_norm,convolution,random_walk.json", "Integrated methods",
    "Fast-Higashi",    "integrated",   "Fast-Higashi",       fast_higashi_root,                                                 "oocyte_zygote_fast_higashi_r%02d",                 "fast_higashi",                                   "fast_higashi.json",                                        "Integrated methods"
  ) %>%
  tidyr::crossing(run = seq_len(n_runs))
) %>%
  mutate(
    seed = seed_start + run - 1L,
    dset_name = sprintf(dset_template, run),
    json_path = purrr::pmap_chr(
      list(roots_chr, dset_name, alg_dir, json_name),
      resolve_json_path
    ),
    exists = !is.na(json_path)
  )

coverage_tbl <- path_tbl %>%
  group_by(run_option, input_label) %>%
  summarise(found = sum(exists), total = n(), .groups = "drop") %>%
  mutate(pct = sprintf("%.1f%%", 100 * found / total))
all_input_status <- path_tbl %>%
  group_by(run_option, input_label, roots_chr, alg_dir, json_name) %>%
  summarise(found = sum(exists), total = n(), .groups = "drop") %>%
  mutate(
    exists = found > 0,
    detail = paste0(found, "/", total, " runs (", sprintf("%.1f%%", 100 * found / total), ")")
  ) %>%
  arrange(run_option, input_label)

print(all_input_status, n = Inf)
```

## 4) Build Unified Per-Run Table

```{r build_unified}
calc_avgbio <- function(df) {
  df %>%
    mutate(avgBio = rowMeans(cbind(best_ari, best_nmi, best_silhouette), na.rm = TRUE))
}

resolve_first_existing <- function(paths) {
  hit <- paths[file.exists(paths)]
  if (length(hit) > 0) hit[[1]] else NA_character_
}

metrics_wide <- path_tbl %>%
  filter(exists) %>%
  rowwise() %>%
  mutate(metrics = list(setNames(lapply(base_metric_cols, function(k) extract_metric_json(json_path, k)), base_metric_cols))) %>%
  tidyr::unnest_wider(metrics) %>%
  ungroup() %>%
  select(run_option, input_code, input_label, family, run, seed, all_of(base_metric_cols))

reshape_external <- function(df, run_option_name) {
  calc_avgbio(df) %>%
    pivot_longer(
      cols = all_of(pivot_metric_cols),
      names_to = "metric",
      values_to = "value"
    ) %>%
    transmute(
      family,
      run_option,
      input_label,
      pipeline_label = paste(run_option_name, "+", input_label),
      run,
      seed,
      metric,
      value
    )
}

reshape_integrated <- function(df, run_option_name) {
  calc_avgbio(df) %>%
    mutate(pipeline_label = run_option_name) %>%
    pivot_longer(
      cols = all_of(pivot_metric_cols),
      names_to = "metric",
      values_to = "value"
    ) %>%
    transmute(
      family,
      run_option,
      input_label,
      pipeline_label,
      run,
      seed,
      metric,
      value
    )
}

inner_df <- metrics_wide %>% filter(run_option == "InnerProduct")
snap_df <- metrics_wide %>% filter(run_option == "SnapATAC_noIDF")
schi_df <- metrics_wide %>% filter(run_option == "scHiCluster")
fast_higashi_df <- metrics_wide %>% filter(run_option == "Fast-Higashi")

long_df <- bind_rows(
  reshape_external(inner_df, "InnerProduct"),
  reshape_external(snap_df, "SnapATAC_noIDF"),
  if (nrow(schi_df) > 0) reshape_integrated(schi_df, "scHiCluster"),
  if (nrow(fast_higashi_df) > 0) reshape_integrated(fast_higashi_df, "Fast-Higashi")
) %>%
  filter(metric %in% metric_order) %>%
  mutate(
    metric = factor(metric, levels = metric_order),
    run_option = factor(run_option, levels = c("InnerProduct", "SnapATAC_noIDF", "scHiCluster", "Fast-Higashi")),
    input_label = factor(input_label, levels = c("Raw", "HiCBZIP-GB(NB)", "HiCBZIP-N(M)", "scHiCImpute", "scHiCluster", "Fast-Higashi")),
    display_group = factor(
      dplyr::case_when(
        run_option == "InnerProduct" ~ "InnerProduct",
        run_option == "SnapATAC_noIDF" ~ "SnapATAC_noIDF",
        TRUE ~ "Integrated methods"
      ),
      levels = c("InnerProduct", "SnapATAC_noIDF", "Integrated methods")
    )
  )

print(long_df, n = 20)
```

## 5) Summary Tables

```{r summary}
summary_df <- long_df %>%
  group_by(run_option, input_label, pipeline_label, metric) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    .groups = "drop"
  ) %>%
  arrange(run_option, input_label, metric)

summary_fmt <- summary_df %>%
  mutate(mean_sd = sprintf("%.3f (%.3f)", mean, sd)) %>%
  select(run_option, input_label, pipeline_label, metric, mean_sd, n)

print(summary_fmt, n = 20)
```

## 6) Main Plot

```{r main_plot, fig.width=16, fig.height=10}
group_boundaries <- tibble(
  metric = levels(long_df$metric),
  idx = seq_along(levels(long_df$metric)),
  metric_group = case_when(
    grepl("^ari|best_ari$", metric, ignore.case = TRUE) ~ "ARI",
    grepl("^nmi|best_nmi$", metric, ignore.case = TRUE) ~ "NMI",
    grepl("silhouette", metric, ignore.case = TRUE) ~ "Silhouette",
    metric == "avgBio" ~ "AvgBIO",
    TRUE ~ "Other"
  )
)

separator_x <- group_boundaries %>%
  group_by(metric_group) %>%
  summarise(last_idx = max(idx), .groups = "drop") %>%
  arrange(last_idx)

group_label_x <- group_boundaries %>%
  group_by(metric_group) %>%
  summarise(
    x = mean(range(idx)),
    .groups = "drop"
  ) %>%
  filter(metric_group %in% c("ARI", "NMI", "Silhouette", "AvgBIO"))

if (nrow(separator_x) > 1) {
  separator_x <- separator_x$last_idx[seq_len(nrow(separator_x) - 1)] + 0.5
} else {
  separator_x <- numeric(0)
}

color_map <- c(
  "Raw" = "#4E79A7",
  "HiCBZIP-GB(NB)" = "#E15759",
  "HiCBZIP-N(M)" = "#59A14F",
  "scHiCImpute" = "#F28E2B",
  "scHiCluster" = "#9C755F",
  "Fast-Higashi" = "#76B7B2"
)

p_main <- ggplot(long_df, aes(x = metric, y = value, fill = input_label, color = input_label)) +
  geom_boxplot(
    width = 0.68,
    alpha = 0.30,
    linewidth = 0.8,
    outlier.shape = NA,
    position = position_dodge(width = 0.78)
  ) +
  geom_point(
    alpha = 0.60,
    size = 1.2,
    position = position_jitterdodge(dodge.width = 0.78, jitter.width = 0.10)
  ) +
  geom_vline(
    xintercept = separator_x,
    linetype = "dashed",
    color = "grey45",
    linewidth = 0.5
  ) +
  geom_text(
    data = group_label_x,
    aes(x = x, y = 1.02, label = metric_group),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 3.2
  ) +
  facet_grid(display_group ~ ., scales = "fixed") +
  scale_fill_manual(values = color_map) +
  scale_color_manual(values = color_map) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Embedding and Clustering Evaluation by Pipeline",
    # subtitle = "InnerProduct and SnapATAC_noIDF compare imputations; scHiCluster is shown as an integrated method",
    x = "Metric",
    y = "Value",
    fill = "Imputation",
    color = "Imputation"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey95"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "top"
  )
```

```{r, fig.width=16, fig.height=10}
print(p_main)
```

## 7) All-Pipeline Plot

```{r all9_plot, fig.width=16, fig.height=13}
pipeline_levels <- c(
  "InnerProduct + Raw",
  "InnerProduct + HiCBZIP-GB(NB)",
  "InnerProduct + HiCBZIP-N(M)",
  "InnerProduct + scHiCImpute",
  "SnapATAC_noIDF + Raw",
  "SnapATAC_noIDF + HiCBZIP-GB(NB)",
  "SnapATAC_noIDF + HiCBZIP-N(M)",
  "SnapATAC_noIDF + scHiCImpute",
  "scHiCluster",
  "Fast-Higashi"
)

plot9_df <- long_df %>%
  mutate(
    pipeline_label = factor(pipeline_label, levels = pipeline_levels),
    metric = factor(metric, levels = metric_order)
  )

pipeline_breaks <- c(4.5, 8.5)
method_label_df <- tibble::tribble(
  ~x,   ~label,
  2.5,  "+InnerProduct",
  6.5,  "+SnapATAC",
  9.5,  "Integrated"
)

p_all9 <- ggplot(plot9_df, aes(x = pipeline_label, y = value, fill = input_label, color = input_label)) +
  geom_boxplot(alpha = 0.30, outlier.shape = NA, width = 0.68) +
  geom_point(alpha = 0.60, size = 1.2, position = position_jitter(width = 0.12)) +
  geom_vline(
    xintercept = pipeline_breaks,
    linetype = "dashed",
    color = "grey45",
    linewidth = 0.5
  ) +
  facet_wrap(~ metric, scales = "fixed", ncol = 6) +
  scale_fill_manual(values = color_map) +
  scale_color_manual(values = color_map) +
  scale_x_discrete(labels = c("Raw", "GB", "NM", "scHiCImpute", "Raw", "GB", "NM", "scHiCImpute", "scHiCluster", "Fast-Higashi")) +
  labs(
    title = "Embedding and Clustering Evaluation by Pipeline",
    # subtitle = "Pipelines are grouped as InnerProduct (first 4), SnapATAC_noIDF (next 4), and scHiCluster (last 1)",
    x = "Pipeline",
    y = "Value",
    fill = "Imputation",
    color = "Imputation"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 35, hjust = 1, vjust = 1),
    strip.background = element_rect(fill = "grey95"),
    legend.position = "top"
  ) +
  geom_text(
    data = method_label_df,
    aes(x = x, y = 1.02, label = label),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 2.5
  ) +
  coord_cartesian(ylim = c(0, 1), clip = "off")
```

```{r, fig.width=16, fig.height=10}
print(p_all9)
```

- scHiCImpute: less reliable for comparison because it retains only 87 cells, because its structural-zero-filtered output (Impute_SZ) produced many extremely sparse or empty contact maps, causing cell loss during chromosome-coverage filtering in SCORE.

- HiCBZIP-N(M): shows the most consistent gain over Raw in both InnerProduct and SnapATAC.

- HiCBZIP-NB(GS): also improves over Raw, but less strongly; gains are clearer in InnerProduct than in SnapATAC.

- SnapATAC: stronger than InnerProduct on this dataset.

- SnapATAC + HiCBZIP-N(M): further improves performance and outperforms scHiCluster, which is one of the strongest integrated methods in this dataset based on the paper.

## 8) Retained Cell Counts

```{r retained_cells}
read_h5ad_nobs <- function(path) {
  if (is.na(path) || !file.exists(path)) return(NA_integer_)
  h5 <- H5File$new(path, mode = "r")
  on.exit(try(h5$close_all(), silent = TRUE), add = TRUE)
  if (!("obs" %in% names(h5))) return(NA_integer_)
  obs_names <- names(h5[["obs"]])
  if (length(obs_names) == 0) return(NA_integer_)
  first_node <- h5[["obs"]][[obs_names[[1]]]]
  if (inherits(first_node, "H5Group") && "codes" %in% names(first_node)) {
    return(length(first_node[["codes"]]$read()))
  }
  vals <- first_node$read()
  if (is.matrix(vals)) return(nrow(vals))
  length(vals)
}

retained_tbl <- tibble::tribble(
  ~run_option,        ~input_label,        ~source_type,   ~source_path,
  "InnerProduct",     "Raw",               "h5ad",         file.path(innerproduct_nb_root, "oocyte_zygote_raw_pairs_r01", "1M", "innerproduct", "anndata_obj.h5ad"),
  "InnerProduct",     "HiCBZIP-GB(NB)",    "h5ad",         file.path(innerproduct_nb_root, "oocyte_zygote_bhzip_pairs_r01", "1M", "innerproduct", "anndata_obj.h5ad"),
  "InnerProduct",     "HiCBZIP-N(M)",      "h5ad",         file.path(innerproduct_nm_root, "oocyte_zygote_bhzip_nm_pairs_r01", "1M", "innerproduct", "anndata_obj.h5ad"),
  "InnerProduct",     "scHiCImpute",       "h5ad",         file.path(innerproduct_schicimpute_root, "oocyte_zygote_schicimpute_pairs_r01", "1M", "innerproduct", "anndata_obj.h5ad"),
  "SnapATAC_noIDF",   "Raw",               "h5ad",         resolve_first_existing(c(
                                              file.path(root_nm, "oocyte_zygote_raw_nm_pairs_snapatac_no_idf_r01", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nm, "oocyte_zygote_raw_nm_pairs_snapatac_no_idf_r01_viz", "1M", "snapatac", "anndata_obj.h5ad")
                                            )),
  "SnapATAC_noIDF",   "HiCBZIP-GB(NB)",    "h5ad",         resolve_first_existing(c(
                                              file.path(root_nb, "oocyte_zygote_bhzip_pairs_snapatac_no_idf_gb_only_r01", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nm, "oocyte_zygote_bhzip_pairs_snapatac_no_idf_gb_only_r01", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nb, "oocyte_zygote_bhzip_pairs_snapatac_no_idf_gb_only_r01_viz", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nm, "oocyte_zygote_bhzip_pairs_snapatac_no_idf_gb_only_r01_viz", "1M", "snapatac", "anndata_obj.h5ad")
                                            )),
  "SnapATAC_noIDF",   "HiCBZIP-N(M)",      "h5ad",         resolve_first_existing(c(
                                              file.path(root_nm, "oocyte_zygote_bhzip_nm_pairs_snapatac_no_idf_r01", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nm, "oocyte_zygote_bhzip_nm_pairs_snapatac_no_idf_r01_viz", "1M", "snapatac", "anndata_obj.h5ad")
                                            )),
  "SnapATAC_noIDF",   "scHiCImpute",       "h5ad",         file.path(snapatac_noidf_schicimpute_root, "oocyte_zygote_schicimpute_pairs_r01", "1M", "snapatac", "anndata_obj.h5ad"),
  "scHiCluster",      "scHiCluster",       "h5ad",         file.path(schicluster_root, "oocyte_zygote_schicluster_r01", "1M", "schicluster%3Avc_sqrt_norm,convolution,random_walk", "anndata_obj.h5ad"),
  "Fast-Higashi",     "Fast-Higashi",      "cluster_json", file.path(fast_higashi_root, "oocyte_zygote_fast_higashi_r01", "1M", "fast_higashi", "clusters_k-means.json")
) %>%
  rowwise() %>%
  mutate(
    retained_cells = if (source_type == "h5ad") {
      read_h5ad_nobs(source_path)
    } else if (source_type == "cluster_json") {
      if (is.na(source_path) || !file.exists(source_path)) {
        NA_integer_
      } else {
        clusters <- jsonlite::fromJSON(source_path, simplifyVector = FALSE)
        length(unique(unlist(clusters, use.names = FALSE)))
      }
    } else {
      NA_integer_
    }
  ) %>%
  ungroup()
```

```{r}
print(retained_tbl %>% select(-source_type, -source_path), n = Inf)
```

## 9) Run-1 PCA2D Visualization

```{r pca2d_plot, fig.width=16, fig.height=11}
read_h5ad_obs_col <- function(obs_group, key) {
  node <- obs_group[[key]]
  if (inherits(node, "H5Group")) {
    nms <- names(node)
    if (all(c("codes", "categories") %in% nms)) {
      codes <- as.integer(node[["codes"]]$read())
      categories <- as.character(node[["categories"]]$read())
      return(as.vector(categories[codes + 1L]))
    }
  }
  if (inherits(node, "H5D")) {
    vals <- node$read()
    attrs <- tryCatch(hdf5r::h5attr_names(node), error = function(e) character(0))
    if ("categories" %in% attrs) {
      cats_obj <- hdf5r::h5attr(node, "categories")
      categories <- as.character(cats_obj$dereference()[[1]]$read())
      return(as.vector(categories[as.integer(vals) + 1L]))
    }
    return(as.vector(as.character(vals)))
  }
  stop("Unsupported obs encoding for key: ", key)
}

read_h5ad_pca2d <- function(path) {
  h5 <- H5File$new(path, mode = "r")
  on.exit(h5$close_all(), add = TRUE)

  x_pca <- as.matrix(h5[["obsm"]][["X_pca"]]$read())
  if (ncol(x_pca) < 2 && nrow(x_pca) >= 2) {
    x_pca <- t(x_pca)
  }
  if (ncol(x_pca) < 2) stop("X_pca has fewer than 2 columns in: ", path)

  obs_cols <- names(h5[["obs"]])
  celltype_key <- if ("celltype" %in% obs_cols) "celltype" else if ("cluster" %in% obs_cols) "cluster" else obs_cols[[1]]
  celltype <- factor(as.vector(read_h5ad_obs_col(h5[["obs"]], celltype_key)))

  if (nrow(x_pca) != length(celltype) && ncol(x_pca) == length(celltype)) {
    x_pca <- t(x_pca)
  }
  if (nrow(x_pca) != length(celltype)) {
    stop("X_pca shape mismatch with celltype labels in: ", path)
  }

  tibble(PC1 = x_pca[, 1], PC2 = x_pca[, 2], celltype = celltype)
}

read_pickle_bundle <- function(path) {
  builtins <- reticulate::import_builtins(convert = FALSE)
  pickle <- reticulate::import("pickle", convert = FALSE)
  handle <- builtins$open(path, "rb")
  on.exit(try(handle$close(), silent = TRUE), add = TRUE)
  obj <- pickle$load(handle)

  z <- as.matrix(reticulate::py_to_r(obj[["z"]]))
  if (ncol(z) < 2 && nrow(z) >= 2) z <- t(z)
  if (ncol(z) < 2) {
    stop("Embedding has fewer than 2 columns in: ", path)
  }

  cells <- as.character(reticulate::py_to_r(obj[["cell"]]))
  if (length(cells) != nrow(z)) {
    stop("Cell names do not match embedding rows in: ", path)
  }

  list(z = z, cell = cells)
}

read_pickle_pca2d <- function(embedding_path, reference_path) {
  bundle <- read_pickle_bundle(embedding_path)
  ref_tbl <- readr::read_tsv(reference_path, show_col_types = FALSE) %>%
    mutate(cell_full = ifelse(grepl("\\.1M$", cell), cell, paste0(cell, ".1M")))

  idx <- match(bundle$cell, ref_tbl$cell_full)
  if (any(is.na(idx))) {
    stop("Some embedding cells are missing from the reference for: ", embedding_path)
  }

  pca <- stats::prcomp(bundle$z, center = TRUE, scale. = FALSE)
  tibble(
    PC1 = pca$x[, 1],
    PC2 = pca$x[, 2],
    celltype = factor(ref_tbl$cluster[idx])
  )
}

reference_path <- "D:/Codex/Test/external/SCORE/examples/data/oocyte_zygote_ref"

pca_map <- tibble::tribble(
  ~run_option,         ~input_label,        ~source_type, ~source_path,                                                                                                           ~cluster_json,
  "InnerProduct",      "Raw",               "h5ad",       file.path(innerproduct_nb_root, "oocyte_zygote_raw_pairs_r01", "1M", "innerproduct", "anndata_obj.h5ad"),              NA_character_,
  "InnerProduct",      "HiCBZIP-GB(NB)",    "h5ad",       file.path(innerproduct_nb_root, "oocyte_zygote_bhzip_pairs_r01", "1M", "innerproduct", "anndata_obj.h5ad"),            NA_character_,
  "InnerProduct",      "HiCBZIP-N(M)",      "h5ad",       file.path(innerproduct_nm_root, "oocyte_zygote_bhzip_nm_pairs_r01", "1M", "innerproduct", "anndata_obj.h5ad"),           NA_character_,
  "InnerProduct",      "scHiCImpute",       "h5ad",       file.path(innerproduct_schicimpute_root, "oocyte_zygote_schicimpute_pairs_r01", "1M", "innerproduct", "anndata_obj.h5ad"), NA_character_,

  "SnapATAC_noIDF",    "Raw",               "h5ad",       resolve_first_existing(c(
                                              file.path(root_nm, "oocyte_zygote_raw_nm_pairs_snapatac_no_idf_r01", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nm, "oocyte_zygote_raw_nm_pairs_snapatac_no_idf_r01_viz", "1M", "snapatac", "anndata_obj.h5ad")
                                            )),                                                                                                                                        NA_character_,
  "SnapATAC_noIDF",    "HiCBZIP-GB(NB)",    "h5ad",       resolve_first_existing(c(
                                              file.path(root_nb, "oocyte_zygote_bhzip_pairs_snapatac_no_idf_gb_only_r01", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nm, "oocyte_zygote_bhzip_pairs_snapatac_no_idf_gb_only_r01", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nb, "oocyte_zygote_bhzip_pairs_snapatac_no_idf_gb_only_r01_viz", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nm, "oocyte_zygote_bhzip_pairs_snapatac_no_idf_gb_only_r01_viz", "1M", "snapatac", "anndata_obj.h5ad")
                                            )),                                                                                                                                        NA_character_,
  "SnapATAC_noIDF",    "HiCBZIP-N(M)",      "h5ad",       resolve_first_existing(c(
                                              file.path(root_nm, "oocyte_zygote_bhzip_nm_pairs_snapatac_no_idf_r01", "1M", "snapatac", "anndata_obj.h5ad"),
                                              file.path(root_nm, "oocyte_zygote_bhzip_nm_pairs_snapatac_no_idf_r01_viz", "1M", "snapatac", "anndata_obj.h5ad")
                                            )),                                                                                                                                        NA_character_,
  "SnapATAC_noIDF",    "scHiCImpute",       "h5ad",       file.path(snapatac_noidf_schicimpute_root, "oocyte_zygote_schicimpute_pairs_r01", "1M", "snapatac", "anndata_obj.h5ad"), NA_character_,

  "scHiCluster",       "scHiCluster",       "h5ad",       file.path(schicluster_root, "oocyte_zygote_schicluster_r01", "1M", "schicluster%3Avc_sqrt_norm,convolution,random_walk", "anndata_obj.h5ad"), NA_character_,
  "Fast-Higashi",      "Fast-Higashi",      "pickle",     file.path(fast_higashi_root, "oocyte_zygote_fast_higashi_r01", "1M", "fast_higashi", "embedding__0.pickle"),               file.path(fast_higashi_root, "oocyte_zygote_fast_higashi_r01", "1M", "fast_higashi", "clusters_k-means.json")
) %>%
  left_join(
    retained_tbl %>% select(run_option, input_label, retained_cells),
    by = c("run_option", "input_label")
  ) %>%
  mutate(
    exists = dplyr::case_when(
      source_type == "h5ad" ~ file.exists(source_path),
      source_type == "pickle" ~ file.exists(source_path),
      TRUE ~ FALSE
    ),
    panel = factor(
      dplyr::case_when(
        run_option == "scHiCluster" ~ paste0("scHiCluster\n(Integrated)\n(n=", retained_cells, ")"),
        run_option == "Fast-Higashi" ~ paste0("Fast-Higashi\n(Integrated)\n(n=", retained_cells, ")"),
        TRUE ~ paste0(run_option, "\n", input_label, "\n", "(n=", retained_cells, ")")
      ),
      levels = c(
        paste0("InnerProduct\nRaw\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "InnerProduct" & retained_tbl$input_label == "Raw"], ")"),
        paste0("InnerProduct\nHiCBZIP-GB(NB)\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "InnerProduct" & retained_tbl$input_label == "HiCBZIP-GB(NB)"], ")"),
        paste0("InnerProduct\nHiCBZIP-N(M)\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "InnerProduct" & retained_tbl$input_label == "HiCBZIP-N(M)"], ")"),
        paste0("InnerProduct\nscHiCImpute\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "InnerProduct" & retained_tbl$input_label == "scHiCImpute"], ")"),
        paste0("SnapATAC_noIDF\nRaw\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "SnapATAC_noIDF" & retained_tbl$input_label == "Raw"], ")"),
        paste0("SnapATAC_noIDF\nHiCBZIP-GB(NB)\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "SnapATAC_noIDF" & retained_tbl$input_label == "HiCBZIP-GB(NB)"], ")"),
        paste0("SnapATAC_noIDF\nHiCBZIP-N(M)\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "SnapATAC_noIDF" & retained_tbl$input_label == "HiCBZIP-N(M)"], ")"),
        paste0("SnapATAC_noIDF\nscHiCImpute\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "SnapATAC_noIDF" & retained_tbl$input_label == "scHiCImpute"], ")"),
        paste0("scHiCluster\n(Integrated)\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "scHiCluster" & retained_tbl$input_label == "scHiCluster"], ")"),
        paste0("Fast-Higashi\n(Integrated)\n", "(n=", retained_tbl$retained_cells[retained_tbl$run_option == "Fast-Higashi" & retained_tbl$input_label == "Fast-Higashi"], ")")
      )
    )
  )

if (!all(pca_map$exists)) {
  warning("Some PCA panels are missing source files; those panels will be blank.")
  print(pca_map %>% filter(!exists) %>% select(run_option, input_label, source_type, source_path, cluster_json))
}

pca_df <- pca_map %>%
  filter(exists) %>%
  mutate(
    pca = purrr::pmap(
      list(source_type, source_path),
      function(source_type, source_path) {
        if (source_type == "h5ad") {
          read_h5ad_pca2d(source_path)
        } else {
          read_pickle_pca2d(source_path, reference_path)
        }
      }
    )
  ) %>%
  tidyr::unnest(pca) %>%
  mutate(panel = factor(panel, levels = levels(pca_map$panel)))

p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = celltype)) +
  geom_point(size = 0.9, alpha = 0.85) +
  facet_wrap(~ panel, ncol = 4, scales = "free", drop = FALSE) +
  labs(
    title = "PCA2D Across Imputation and Embedding Pipelines",
    # subtitle = "Missing combinations are left blank when no anndata_obj.h5ad is available",
    x = "PC1",
    y = "PC2",
    color = "Celltype"
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey95"),
    legend.position = "right"
  )
```


```{r, fig.width=15, fig.height=10}
print(p_pca)
```

## 10) Example Contact Heatmap + scHiCluster Scaled

```{r example_heatmap_with_schicluster, fig.width=20, fig.height=5.6}
anchor_path <- "D:/Codex/Test/external/SCORE/examples/data/mm10.genome_split_1M"
raw_pair_dir <- "D:/Codex/Test/external/SCORE/examples/data/oocyte_zygote_mm10/1M"
nb_pair_dir  <- "D:/Codex/Test/data/bhzip_score_compare_from_pairs/bhzip_1M"
nm_pair_dir  <- "D:/Codex/Test/data/bhzip_score_compare_from_pairs_nm/bhzip_nm_1M"
schic_pair_dir <- "D:/Codex/Test/data/schicimpute_score_compare_from_pairs/schicimpute_1M"
schi_cap_quantile <- 0.96

target_chr <- "chr1"
requested_cell <- NULL

anchors <- read.delim(anchor_path, sep = "\t", header = FALSE, stringsAsFactors = FALSE)
colnames(anchors) <- c("chr", "start", "end", "idx0")
chr_ids <- anchors$idx0[anchors$chr == target_chr]

read_pair_file <- function(path) {
  if (!file.exists(path) || file.info(path)$size == 0) return(NULL)
  x <- tryCatch(read.table(path, sep = "\t", header = FALSE, stringsAsFactors = FALSE), error = function(e) NULL)
  if (is.null(x) || nrow(x) == 0 || ncol(x) < 3) return(NULL)
  x <- x[, 1:3]
  colnames(x) <- c("bin1", "bin2", "count")
  x$bin1 <- suppressWarnings(as.integer(x$bin1))
  x$bin2 <- suppressWarnings(as.integer(x$bin2))
  x$count <- suppressWarnings(as.numeric(x$count))
  x <- x[!is.na(x$bin1) & !is.na(x$bin2) & !is.na(x$count), , drop = FALSE]
  if (nrow(x) == 0) return(NULL)
  x
}

build_chr_mat <- function(pair_dir, cell_file, chr_ids) {
  n_bins <- length(chr_ids)
  f <- file.path(pair_dir, cell_file)
  d <- read_pair_file(f)
  if (is.null(d)) return(matrix(0, nrow = n_bins, ncol = n_bins))

  keep <- d$bin1 %in% chr_ids & d$bin2 %in% chr_ids
  d <- d[keep, , drop = FALSE]
  if (nrow(d) == 0) return(matrix(0, nrow = n_bins, ncol = n_bins))

  old2new <- integer(max(chr_ids) + 1L)
  old2new[chr_ids + 1L] <- seq_len(n_bins)
  i <- old2new[d$bin1 + 1L]
  j <- old2new[d$bin2 + 1L]
  ok <- i > 0 & j > 0 & !is.na(i) & !is.na(j)
  i <- i[ok]; j <- j[ok]; w <- d$count[ok]
  if (length(i) == 0) return(matrix(0, nrow = n_bins, ncol = n_bins))

  mat <- matrix(0, nrow = n_bins, ncol = n_bins)
  for (k in seq_along(i)) {
    ii <- i[k]; jj <- j[k]
    mat[ii, jj] <- mat[ii, jj] + w[k]
    if (ii != jj) mat[jj, ii] <- mat[jj, ii] + w[k]
  }
  mat
}

pick_example_cell <- function(dirs) {
  available <- dirs[dir.exists(dirs)]
  if (length(available) == 0) return(NA_character_)
  file_sets <- lapply(available, function(d) list.files(d, full.names = FALSE))
  common <- Reduce(intersect, file_sets)
  if (length(common) == 0) return(NA_character_)
  if (!is.null(requested_cell) && requested_cell %in% common) return(requested_cell)
  sort(common)[1]
}

cell_file <- pick_example_cell(c(raw_pair_dir, nb_pair_dir, nm_pair_dir, schic_pair_dir))

vc_sqrt_norm <- function(mat) {
  sm <- colSums(mat)
  sm[sm == 0] <- 1
  sm <- sqrt(sm)
  denom <- tcrossprod(sm, sm)
  mat / denom
}

convolution_mean <- function(mat, kernel_shape = 3L) {
  stopifnot(kernel_shape %% 2L == 1L)
  pad <- 1L
  nr <- nrow(mat)
  nc <- ncol(mat)
  padded <- matrix(0, nrow = nr + 2L * pad, ncol = nc + 2L * pad)
  padded[(pad + 1L):(pad + nr), (pad + 1L):(pad + nc)] <- mat
  out <- matrix(0, nrow = nr, ncol = nc)
  for (i in seq_len(nr)) {
    for (j in seq_len(nc)) {
      out[i, j] <- mean(padded[i:(i + 2L * pad), j:(j + 2L * pad)])
    }
  }
  out
}

random_walk_smooth <- function(mat, random_walk_ratio = 1.0, t = 1L) {
  sm <- rowSums(mat)
  sm[sm == 0] <- 1
  walk <- mat / sm
  out <- mat
  for (step in seq_len(t)) {
    out <- random_walk_ratio * (out %*% walk) + (1 - random_walk_ratio) * out
  }
  out
}

input_tbl <- tibble::tribble(
  ~method,            ~pair_dir,       ~label,
  "Raw",              raw_pair_dir,    "Raw",
  "HiCBZIP-GB(NB)",   nb_pair_dir,     "HiCBZIP-GB(NB)",
  "HiCBZIP-N(M)",     nm_pair_dir,     "HiCBZIP-N(M)",
  "scHiCImpute",      schic_pair_dir,  "scHiCImpute"
)

mat_to_df <- function(mat, method, label, available, note) {
  if (!available || is.null(mat)) {
    return(tibble::tibble(
      method = method, label = label, i = 1, j = 1, count = NA_real_,
      available = FALSE, note = note
    ))
  }
  n <- nrow(mat)
  expand.grid(i = seq_len(n), j = seq_len(n), KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE) %>%
    mutate(
      count = as.numeric(mat[cbind(i, j)]),
      method = method,
      label = label,
      available = TRUE,
      note = ""
    )
}

hm_list <- lapply(seq_len(nrow(input_tbl)), function(ix) {
  method_i <- input_tbl$method[ix]
  dir_i <- input_tbl$pair_dir[ix]
  label_i <- input_tbl$label[ix]

  if (!dir.exists(dir_i)) {
    return(mat_to_df(NULL, method_i, label_i, FALSE, paste("Missing dir:", dir_i)))
  }
  if (is.na(cell_file) || !file.exists(file.path(dir_i, cell_file))) {
    return(mat_to_df(NULL, method_i, label_i, FALSE, paste("Missing cell file:", cell_file)))
  }
  mat <- build_chr_mat(dir_i, cell_file, chr_ids)
  mat_to_df(mat, method_i, label_i, TRUE, "")
})

raw_mat <- build_chr_mat(raw_pair_dir, cell_file, chr_ids)
schicluster_mat <- raw_mat %>%
  vc_sqrt_norm() %>%
  convolution_mean(kernel_shape = 3L) %>%
  random_walk_smooth(random_walk_ratio = 1.0, t = 1L)

heat_df <- bind_rows(hm_list) %>%
  mutate(
    method = factor(method, levels = c("Raw", "HiCBZIP-GB(NB)", "HiCBZIP-N(M)", "scHiCImpute")),
    value = log1p(count)
  )

note_df <- heat_df %>%
  group_by(method, label, available, note) %>%
  summarise(.groups = "drop") %>%
  filter(!available)

pos_vals <- heat_df$value[is.finite(heat_df$value) & heat_df$value > 0]
count_max <- if (length(pos_vals) > 0) {
  as.numeric(stats::quantile(pos_vals, 0.995, na.rm = TRUE))
} else {
  suppressWarnings(max(heat_df$value, na.rm = TRUE))
}
if (!is.finite(count_max)) count_max <- 1
count_max <- max(count_max, 1e-6)

schi_df <- expand.grid(
  i = seq_len(nrow(schicluster_mat)),
  j = seq_len(ncol(schicluster_mat)),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
) %>%
  mutate(
    value = as.numeric(schicluster_mat[cbind(i, j)]),
    log_value = log1p(value)
  )

schi_cap <- as.numeric(stats::quantile(schi_df$log_value[is.finite(schi_df$log_value)], schi_cap_quantile, na.rm = TRUE))
if (!is.finite(schi_cap)) schi_cap <- 1

schi_df <- schi_df %>%
  mutate(
    scaled_value = pmin(log_value / schi_cap, 1) * count_max,
    method = factor("scHiCluster (scaled)",
      levels = c("Raw", "HiCBZIP-GB(NB)", "HiCBZIP-N(M)", "scHiCImpute", "scHiCluster (scaled)")
    ),
    available = TRUE,
    note = ""
  )

heat_df_plot <- heat_df %>%
  mutate(
    method = factor(as.character(method),
      levels = c("Raw", "HiCBZIP-GB(NB)", "HiCBZIP-N(M)", "scHiCImpute", "scHiCluster (scaled)")
    ),
    plot_value = value
  ) %>%
  select(method, i, j, plot_value, available, note) %>%
  bind_rows(
    schi_df %>%
      transmute(method, i, j, plot_value = scaled_value, available, note)
  )

note_df <- heat_df_plot %>%
  group_by(method, available, note) %>%
  summarise(.groups = "drop") %>%
  filter(!available)

p_heat <- ggplot(heat_df_plot, aes(x = j, y = i, fill = plot_value)) +
  geom_tile() +
  facet_wrap(~ method, ncol = 5, drop = FALSE) +
  scale_fill_viridis_c(
    limits = c(0, count_max),
    option = "plasma",
    oob = scales::squish,
    na.value = "grey95",
    name = "scaled\nlog1p"
  ) +
  coord_equal() +
  scale_y_reverse() +
  labs(
    title = paste0("Example Contact Heatmap | ", target_chr, " | cell: ", cell_file),
    subtitle = "Raw, HiCBZIP, and scHiCImpute use log1p(count); scHiCluster is shown as scaled transformed output",
    x = "bin",
    y = "bin"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey95"),
    strip.text = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "right"
  )

if (nrow(note_df) > 0) {
  p_heat <- p_heat +
    geom_text(
      data = note_df,
      aes(x = 1, y = 1, label = note),
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 1,
      size = 3.2
    )
}
```


```{r, fig.width=20, fig.height=5}
print(p_heat)
```

## 11) Save Outputs

```{r save}
out_dir <- "D:/Codex/Test/downloaded_results/paper_metrics_external_impute_plus_integrated_with_fast_higashi"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(long_df, file.path(out_dir, "per_run_metrics_long.csv"))
readr::write_csv(summary_df, file.path(out_dir, "summary_mean_sd.csv"))
readr::write_csv(pca_map, file.path(out_dir, "pca_h5ad_resolution_table.csv"))
readr::write_csv(retained_tbl, file.path(out_dir, "retained_cells_run1.csv"))

ggsave(
  filename = file.path(out_dir, "external_impute_plus_integrated_with_fast_higashi_main.png"),
  plot = p_main,
  width = 14,
  height = 11,
  dpi = 220
)

ggsave(
  filename = file.path(out_dir, "external_impute_plus_integrated_with_fast_higashi_all_pipelines.png"),
  plot = p_all9,
  width = 16,
  height = 13,
  dpi = 220
)

ggsave(
  filename = file.path(out_dir, "external_impute_plus_integrated_with_fast_higashi_pca2d.png"),
  plot = p_pca,
  width = 15,
  height = 10,
  dpi = 220
)

ggsave(
  filename = file.path(out_dir, "external_impute_plus_integrated_with_fast_higashi_example_heatmap_with_schicluster.png"),
  plot = p_heat,
  width = 20,
  height = 5.8,
  dpi = 220
)

cat("Saved outputs to:", out_dir, "\n")
```
