---
title: "Flayamer InnerProduct scHiCTools Parity (1M)"
author: "Codex + User"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE, fig.width = 12, fig.height = 6)
set.seed(12345)
```

## 1) Libraries and Paths

```{r libs}
required_pkgs <- c(
  "Matrix", "dplyr", "readr", "tibble", "ggplot2",
  "uwot", "mclust", "tidyr", "purrr", "jsonlite"
)

missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop(
    "Missing packages: ", paste(missing_pkgs, collapse = ", "),
    "\nInstall them first, e.g. install.packages(missing_pkgs)."
  )
}

library(Matrix)
library(dplyr)
library(readr)
library(tibble)
library(ggplot2)
library(uwot)
library(mclust)
library(tidyr)
library(purrr)
library(jsonlite)

proj_dir <- normalizePath(".", winslash = "/", mustWork = TRUE)
data_root <- file.path(proj_dir, "data", "raw", "flayamer")
out_dir <- file.path(proj_dir, "outputs", "flayamer_innerproduct_schictools_parity")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

source(file.path(proj_dir, "R", "helpers_ramani_io.R"))
source(file.path(proj_dir, "R", "helpers_embedding.R"))
source(file.path(proj_dir, "previous_R_code_example", "BHZIP_EB_250911.R"))
source(file.path(proj_dir, "previous_R_code_example", "Functions.R"))

cat("Project dir:", proj_dir, "\n")
cat("Data root:", data_root, "\n")
cat("Output dir:", out_dir, "\n")

if (!dir.exists(data_root)) {
  stop("Flayamer data directory not found: ", data_root)
}
```

## 2) Config

```{r config}
cfg <- list(
  seed = 12345,
  run_full = TRUE,
  matrix_file_override = file.path(data_root, "GSE305523_reanalysis_GSE80006_oocyte_zygote", "oocyte_zygote", "oocyte_zygote_1M.scool"),
  metadata_file_override = NA_character_,
  target_chromosomes = NA,
  # QC
  qc_min_depth = 5000,
  qc_depth_col = "depth",
  qc_cis_min_contacts = NULL,
  # Imputation (disabled for raw parity run)
  run_imputation = FALSE,
  bzip_r = 1,
  bzip_threshold = 3,
  bzip_B = 10,
  bzip_include_diag = TRUE,
  use_imputation_cache = FALSE,
  # Embedding (InnerProduct only)
  embedding_pipelines = c(
    "paper_innerproduct_mds_fixed",
    "paper_innerproduct_mds_1M",
    "paper_innerproduct_mds_2M",
    "paper_innerproduct_mds_4M",
    "paper_innerproduct_mds_6M",
    "paper_innerproduct_mds_10M",
    "paper_innerproduct_mds_15M",
    "paper_innerproduct_mds_20M",
    "paper_innerproduct_mds_30M",
    "paper_innerproduct_mds_50M",
    "paper_innerproduct_mds_n10",
    "paper_innerproduct_mds_n11",
    "paper_innerproduct_mds_n32"
  ),
  innerproduct_max_ranges_mb = c(1, 2, 4, 6, 10, 15, 20, 30, 50),
  innerproduct_n_strata = 10L,
  schi_pcs_per_chr = 10,
  npc = 10,
  k_nstart = 50,
  # Fair/common robustness for evaluation
  embedding_eval_cov_q = c(0, 1),
  run_heatmap = FALSE,
  # Additional paper-style chromosome QC (minimum cis-contacts per chromosome)
  qc_chr_min_contacts_by_chr_mb = TRUE,
  qc_chr_exclude = c("chrX", "chrY"),
  # scHiCTools python reference branch (no SCORE CLI)
  run_schictools_reference = TRUE,
  schictools_out_dir = file.path(out_dir, "schictools_reference"),
  schictools_python = "C:/ProgramData/Anaconda3/python.exe"
)
set.seed(cfg$seed)
```

## 3) Detect Flayamer Files

```{r detect_files}
all_files <- list.files(data_root, recursive = TRUE, full.names = TRUE)
if (length(all_files) == 0) stop("No files found under ", data_root)

meta_candidates <- all_files[
  grepl("\\.(tsv|tsv\\.gz|csv|csv\\.gz)$", tolower(all_files)) &
    grepl("meta|metadata|label|annot", tolower(basename(all_files)))
]

metadata_file <- if (!is.na(cfg$metadata_file_override)) {
  cfg$metadata_file_override
} else if (length(meta_candidates) > 0) {
  sort(unique(meta_candidates))[[1]]
} else {
  stop("Could not auto-detect metadata file in ", data_root,
       ". Set cfg$metadata_file_override.")
}

candidates <- list_candidate_matrix_files(data_root)
candidates <- setdiff(candidates, metadata_file)
if (length(candidates) == 0) {
  stop("Could not auto-detect matrix-like file in ", data_root,
       ". Set cfg$matrix_file_override.")
}

matrix_file <- if (!is.na(cfg$matrix_file_override)) cfg$matrix_file_override else pick_default_matrix_file(candidates)
sidecar <- find_sidecar_ids(matrix_file)

cat("Metadata file:\n", metadata_file, "\n")
cat("Matrix file:\n", matrix_file, "\n")
print(sidecar)
```

## 4) Load Matrix and Align Metadata

```{r load_matrix}
read_meta_auto <- function(path) {
  lower <- tolower(path)
  if (grepl("\\.csv(\\.gz)?$", lower)) {
    readr::read_csv(path, show_col_types = FALSE, progress = FALSE)
  } else {
    readr::read_tsv(path, show_col_types = FALSE, progress = FALSE)
  }
}

meta <- read_meta_auto(metadata_file)
meta_cols <- infer_metadata_columns(meta)
cell_col <- meta_cols$cell_col
label_col <- meta_cols$label_col

cat("Detected metadata cell column:", cell_col, "\n")
cat("Detected metadata label column:", label_col, "\n")

chunked_allchr_mode <- FALSE
selected_chromosomes <- NULL

if (grepl("\\.scool$", tolower(matrix_file))) {
  available_cells <- list_scool_cells(matrix_file)
  # Flayamer/GSE80006 metadata IDs omit the resolution suffix present in .scool IDs (e.g. ".1M").
  normalize_flayamer_key <- function(x) {
    x <- as.character(x)
    x <- gsub("\\.1m$", "", x, ignore.case = TRUE)
    tolower(x)
  }

  meta0 <- meta %>%
    dplyr::filter(!is.na(.data[[cell_col]]), !is.na(.data[[label_col]])) %>%
    dplyr::mutate(.match_key = normalize_flayamer_key(.data[[cell_col]]))

  scool_tbl <- tibble::tibble(
    .scool_cell = available_cells,
    .match_key = normalize_flayamer_key(available_cells)
  )

  direct_key_map <- meta0 %>%
    dplyr::inner_join(scool_tbl, by = ".match_key") %>%
    dplyr::mutate(.match_method = "flayamer_key")
  cat("Flayamer key-match cells:", nrow(direct_key_map), "\n")

  if (nrow(direct_key_map) >= 50) {
    meta_all <- direct_key_map
    meta_all[[cell_col]] <- meta_all$.scool_cell
  } else {
    meta_all <- match_metadata_to_scool_cells(
      meta_df = meta,
      cell_col = cell_col,
      label_col = label_col,
      scool_cells = available_cells,
      min_cells = 50
    )
  }

  if (length(cfg$target_chromosomes) == 1 && is.na(cfg$target_chromosomes)) {
    selected_chromosomes <- list_scool_chromosomes(matrix_file)
    selected_chromosomes <- setdiff(selected_chromosomes, c("chrM", "MT", "M"))
    chunked_allchr_mode <- TRUE
    Y_all <- NULL
  } else {
    selected_chromosomes <- cfg$target_chromosomes
    Y_all <- load_scool_feature_by_cell(
      scool_path = matrix_file,
      cell_ids = meta_all[[cell_col]],
      target_chromosomes = selected_chromosomes,
      include_diag = cfg$bzip_include_diag,
      verbose = TRUE
    )
    meta_all <- meta_all %>% arrange(match(.data[[cell_col]], colnames(Y_all)))
  }
} else {
  mx <- read_contact_matrix(
    matrix_path = matrix_file,
    cell_ids_path = sidecar$cell_ids_path,
    feature_ids_path = sidecar$feature_ids_path
  )

  feature_by_cell <- as_feature_by_cell(
    mat = mx$mat,
    cell_ids = mx$cell_ids,
    metadata_cells = meta[[cell_col]]
  )

  aligned <- align_matrix_and_metadata(
    feature_by_cell = feature_by_cell,
    meta_df = meta,
    cell_col = cell_col,
    label_col = label_col
  )

  Y_all <- aligned$mat
  meta_all <- aligned$meta
  if (inherits(Y_all, "Matrix")) {
    Y_all <- as(Y_all, "dgCMatrix")
  } else {
    Y_all <- as.matrix(Y_all)
  }
}

meta_all <- meta_all %>%
  group_by(.data[[label_col]]) %>%
  filter(n() >= 20) %>%
  ungroup()

# Paper-style QC before imputation/embedding: remove very low-depth cells.
if (!is.null(cfg$qc_min_depth) && is.finite(cfg$qc_min_depth) && cfg$qc_min_depth > 0) {
  depth_col <- cfg$qc_depth_col
  if (!is.null(depth_col) && depth_col %in% names(meta_all)) {
    n_before <- nrow(meta_all)
    meta_all <- meta_all %>% filter(.data[[depth_col]] >= cfg$qc_min_depth)
    cat("QC depth filter (", depth_col, " >= ", cfg$qc_min_depth, "): kept ",
        nrow(meta_all), " / ", n_before, " cells\n", sep = "")
  } else {
    stop("SCORE-style QC requires metadata depth column. Missing: ", depth_col)
  }
}

if (!is.null(Y_all)) {
  Y_all <- Y_all[, meta_all[[cell_col]], drop = FALSE]
  if (!all(meta_all[[cell_col]] == colnames(Y_all))) stop("Cell ordering mismatch after alignment.")

  cat("Aligned matrix dim (features x cells):", nrow(Y_all), "x", ncol(Y_all), "\n")
  cat("Aligned metadata rows:", nrow(meta_all), "\n")
} else {
  cat("Using chunked all-chromosome mode. Metadata rows:", nrow(meta_all), "\n")
}
```

## 5) Build Full Benchmark Matrix

```{r full_matrix}
if (isTRUE(chunked_allchr_mode)) {
  build_scool_cells_by_chr <- function(cell_ids, chromosomes, verbose = TRUE) {
    mats <- lapply(seq_along(chromosomes), function(i) {
      chr <- chromosomes[[i]]
      if (verbose) cat("Loading chromosome", chr, "(", i, "/", length(chromosomes), ")\n")
      m <- load_scool_feature_by_cell(
        scool_path = matrix_file,
        cell_ids = cell_ids,
        target_chromosomes = chr,
        include_diag = cfg$bzip_include_diag,
        verbose = FALSE
      )
      rownames(m) <- paste0(chr, "::", rownames(m))
      m
    })
    do.call(rbind, mats)
  }

  cells_full <- meta_all[[cell_col]]
  Y_full <- build_scool_cells_by_chr(cells_full, selected_chromosomes, verbose = TRUE)
} else {
  Y_full <- Y_all
}

labels_full <- meta_all[[label_col]]
names(labels_full) <- meta_all[[cell_col]]

# Additional chromosome-level QC, applied before imputation/embedding.
if (isTRUE(cfg$qc_chr_min_contacts_by_chr_mb)) {
  infer_resolution_mb_from_path <- function(path) {
    b <- tolower(basename(path))
    m1 <- regexec("([0-9]+)m\\.scool$", b)
    g1 <- regmatches(b, m1)[[1]]
    if (length(g1) == 2) return(as.numeric(g1[2]))
    m2 <- regexec("([0-9]+)kb\\.scool$", b)
    g2 <- regmatches(b, m2)[[1]]
    if (length(g2) == 2) return(as.numeric(g2[2]) / 1000)
    1
  }

  feat_names <- rownames(Y_full)
  has_chr_prefix <- !is.null(feat_names) && any(grepl("::", feat_names, fixed = TRUE))
  if (has_chr_prefix) {
    res_mb <- infer_resolution_mb_from_path(matrix_file)
    chr_vec <- sub("::.*$", "", feat_names)
    chr_groups <- split(seq_len(nrow(Y_full)), chr_vec)
    chr_use <- setdiff(names(chr_groups), as.character(cfg$qc_chr_exclude))

    keep_cells <- rep(TRUE, ncol(Y_full))
    names(keep_cells) <- colnames(Y_full)
    fail_summary <- vector("list", length(chr_groups))
    names(fail_summary) <- names(chr_groups)

    for (chr in chr_use) {
      idx <- chr_groups[[chr]]
      n_bins <- infer_n_from_pairs(length(idx))
      chr_len_mb <- n_bins * res_mb
      min_contacts_chr <- max(1, as.integer(chr_len_mb))
      chr_contacts <- Matrix::colSums(Y_full[idx, , drop = FALSE])
      pass_chr <- chr_contacts >= min_contacts_chr
      keep_cells <- keep_cells & pass_chr
      pass_rate <- mean(pass_chr)
      fail_summary[[chr]] <- c(
        min_contacts = min_contacts_chr,
        pass_rate = pass_rate,
        n_fail = sum(!pass_chr)
      )
      cat("Chr-QC ", chr, ": min=", min_contacts_chr, ", pass_rate=", round(pass_rate, 3), "\n", sep = "")
    }

    fail_tbl <- dplyr::bind_rows(lapply(chr_use, function(chr) {
      x <- fail_summary[[chr]]
      tibble::tibble(chr = chr, min_contacts = x[["min_contacts"]], pass_rate = x[["pass_rate"]], n_fail = x[["n_fail"]])
    })) %>%
      dplyr::arrange(pass_rate, dplyr::desc(n_fail))
    print(fail_tbl)

    n_before <- ncol(Y_full)
    kept_ids <- names(keep_cells)[keep_cells]
    Y_full <- Y_full[, kept_ids, drop = FALSE]
    meta_all <- meta_all %>% dplyr::filter(.data[[cell_col]] %in% kept_ids) %>%
      dplyr::arrange(match(.data[[cell_col]], kept_ids))
    labels_full <- meta_all[[label_col]]
    names(labels_full) <- meta_all[[cell_col]]
    cat("QC chromosome-level filter: kept ", ncol(Y_full), " / ", n_before, " cells\n", sep = "")
    if (ncol(Y_full) == 0) {
      stop("Chromosome-level QC removed all cells. Inspect fail table above and verify chromosome coverage rule on this dataset.")
    }
  } else {
    warning("Skipping chromosome-level QC: features are not chr::pair labeled.")
  }
}

cat("Full matrix dim:", nrow(Y_full), "x", ncol(Y_full), "\n")
print(table(labels_full))
```

## 6) Coverage (Raw Parity)

```{r imputation}
cov_full <- compute_bulk_referenced_coverage(Y_full)

validate_bzip_input <- function(Y, coverage, include_diag_expected = TRUE) {
  if (!is.matrix(Y)) Y <- as.matrix(Y)
  if (ncol(Y) != length(coverage)) {
    stop("coverage length mismatch: ncol(Y)=", ncol(Y), " vs length(coverage)=", length(coverage))
  }

  feat_names <- rownames(Y)
  has_chr_prefix <- !is.null(feat_names) && any(grepl("::", feat_names, fixed = TRUE))

  check_one <- function(n_pairs, tag) {
    n <- infer_n_from_long(n_pairs)
    message("Validated ", tag, ": n_pairs=", n_pairs, " -> n_bins=", n)
  }

  if (has_chr_prefix) {
    chr_vec <- sub("::.*$", "", feat_names)
    idx_list <- split(seq_len(nrow(Y)), chr_vec)
    for (chr in names(idx_list)) check_one(length(idx_list[[chr]]), paste0("chr ", chr))
  } else {
    check_one(nrow(Y), "single-matrix")
  }

  if (!isTRUE(include_diag_expected)) {
    warning("include_diag=FALSE not supported for .scool loader path in this notebook")
  }
  invisible(TRUE)
}

run_one_method <- function(Y, coverage, method_name = "HiCBZIP-GB(NB)") {
  message("Running ", method_name, " with ", ncol(Y), " cells and ", nrow(Y), " features")

  feat_names <- rownames(Y)
  has_chr_prefix <- !is.null(feat_names) && any(grepl("::", feat_names, fixed = TRUE))

  if (!has_chr_prefix) {
    return(run_BZIP_GB_NB(
      sim_y = as.matrix(Y),
      coverage = coverage,
      r = cfg$bzip_r,
      threshold = cfg$bzip_threshold,
      B = cfg$bzip_B,
      include_diag = cfg$bzip_include_diag
    ))
  }

  chr_vec <- sub("::.*$", "", feat_names)
  idx_list <- split(seq_len(nrow(Y)), chr_vec)

  out <- matrix(0, nrow = nrow(Y), ncol = ncol(Y))
  rownames(out) <- rownames(Y)
  colnames(out) <- colnames(Y)

  for (chr in names(idx_list)) {
    idx <- idx_list[[chr]]
    Yc <- Y[idx, , drop = FALSE]

    ok <- TRUE
    tryCatch({
      infer_n_from_long(nrow(Yc))
    }, error = function(e) {
      ok <<- FALSE
      warning("Skipping chromosome ", chr, " because N_pairs is not triangular: ", nrow(Yc))
    })
    if (!ok) next

    out[idx, ] <- run_BZIP_GB_NB(
      sim_y = as.matrix(Yc),
      coverage = coverage,
      r = cfg$bzip_r,
      threshold = cfg$bzip_threshold,
      B = cfg$bzip_B,
      include_diag = cfg$bzip_include_diag
    )
  }

  out
}

imputed_full <- NULL
if (isTRUE(cfg$run_full) && isTRUE(cfg$run_imputation)) {
  cache_tag <- paste0("imputed_flayamer_", ncol(Y_full), "cells_", nrow(Y_full), "features.rds")
  cache_path <- file.path(out_dir, cache_tag)

  if (isTRUE(cfg$use_imputation_cache) && file.exists(cache_path)) {
    cat("Loading cached imputation:", cache_path, "\n")
    imputed_full <- readRDS(cache_path)
  } else {
    validate_bzip_input(Y_full, cov_full$coverage, include_diag_expected = cfg$bzip_include_diag)
    imputed_full <- run_one_method(Y_full, cov_full$coverage)
    if (isTRUE(cfg$use_imputation_cache)) {
      saveRDS(imputed_full, cache_path)
      cat("Saved imputation cache:", cache_path, "\n")
    }
  }
  cat("Imputed dim:", dim(imputed_full)[1], "x", dim(imputed_full)[2], "\n")
} else {
  cat("Imputation skipped (cfg$run_imputation=FALSE); Raw-only parity mode.\n")
}
```

## 6b) Example Contact Heatmap (Raw vs Imputed)

```{r example_heatmap, fig.height=6}
vec_upper_rowmajor_to_mat_simple <- function(v) {
  n <- infer_n_from_long(length(v))
  M <- matrix(0, n, n)
  idx <- 1L
  for (i in seq_len(n)) {
    len <- n - i + 1L
    M[i, i:n] <- v[idx:(idx + len - 1L)]
    idx <- idx + len
  }
  M[lower.tri(M)] <- t(M)[lower.tri(M)]
  M
}

to_long <- function(M, method_name) {
  df <- as.data.frame(as.table(M))
  colnames(df) <- c("i", "j", "value")
  df$i <- as.integer(df$i)
  df$j <- as.integer(df$j)
  df$value <- as.numeric(df$value)
  df <- df[is.finite(df$i) & is.finite(df$j) & is.finite(df$value), , drop = FALSE]
  df$method <- method_name
  df
}

if (!isTRUE(cfg$run_heatmap)) {
  cat("Heatmap skipped: cfg$run_heatmap=FALSE\n")
} else if (is.null(rownames(Y_full)) || !any(grepl("::", rownames(Y_full), fixed = TRUE))) {
  cat("Heatmap skipped: chromosome-tagged rownames (chr::pair) not found.\n")
} else {
  chr_vec <- sub("::.*$", "", rownames(Y_full))

  # Pick a representative higher-coverage cell to avoid empty visuals.
  cell_idx <- order(cov_full$coverage, decreasing = TRUE)[max(1, floor(0.1 * length(cov_full$coverage)))]
  cell_name <- colnames(Y_full)[cell_idx]

  # Choose chromosome with most nonzero counts in this cell.
  chr_nonzero <- tapply(Y_full[, cell_idx] > 0, chr_vec, sum)
  chr_show <- names(which.max(chr_nonzero))
  idx_chr <- which(chr_vec == chr_show)

  raw_vec <- as.numeric(Y_full[idx_chr, cell_idx])
  imp_vec <- as.numeric(imputed_full[idx_chr, cell_idx])

  raw_mat <- vec_upper_rowmajor_to_mat_simple(raw_vec)
  imp_mat <- vec_upper_rowmajor_to_mat_simple(imp_vec)
  diff_mat <- imp_mat - raw_mat

  nz <- which((raw_mat + imp_mat) > 0, arr.ind = TRUE)
  if (nrow(nz) > 0) {
    pad <- 8
    i_min <- max(1, min(nz[, 1], nz[, 2]) - pad)
    i_max <- min(nrow(raw_mat), max(nz[, 1], nz[, 2]) + pad)
    raw_mat <- raw_mat[i_min:i_max, i_min:i_max, drop = FALSE]
    imp_mat <- imp_mat[i_min:i_max, i_min:i_max, drop = FALSE]
    diff_mat <- diff_mat[i_min:i_max, i_min:i_max, drop = FALSE]
  }

  raw_log <- log1p(raw_mat)
  imp_log <- log1p(imp_mat)
  pos_vals <- c(raw_log[raw_log > 0], imp_log[imp_log > 0])
  count_max <- if (length(pos_vals) > 0) as.numeric(stats::quantile(pos_vals, 0.995, na.rm = TRUE)) else max(c(raw_log, imp_log), na.rm = TRUE)
  count_max <- max(count_max, 1e-6)

  heat_df <- bind_rows(
    to_long(raw_log, "Raw"),
    to_long(imp_log, "HiCBZIP-GB(NB)")
  )

  p_hm <- ggplot(heat_df, aes(x = j, y = i, fill = value)) +
    geom_tile() +
    facet_wrap(~method, ncol = 2) +
    scale_fill_viridis_c(limits = c(0, count_max), option = "plasma") +
    coord_equal() +
    scale_y_reverse() +
    theme_minimal() +
    labs(
      title = paste0("Flayamer Example Contact Heatmap on ", chr_show, " (zoomed)"),
      subtitle = paste0("Cell: ", cell_name, " | scale=log1p(count), cap=99.5%"),
      x = "bin", y = "bin", fill = "log1p"
    )

  print(p_hm)

  diff_df <- to_long(diff_mat, "Imputed - Raw")
  lim <- max(abs(diff_df$value), na.rm = TRUE)
  lim <- ifelse(is.finite(lim) && lim > 0, lim, 1)

  p_diff <- ggplot(diff_df, aes(x = j, y = i, fill = value)) +
    geom_tile() +
    scale_fill_gradient2(low = "navy", mid = "white", high = "firebrick", midpoint = 0, limits = c(-lim, lim)) +
    coord_equal() +
    scale_y_reverse() +
    theme_minimal() +
    labs(
      title = paste0("Flayamer Difference Map on ", chr_show),
      subtitle = "Imputed - Raw",
      x = "bin", y = "bin", fill = "delta"
    )

  print(p_diff)

  ggsave(file.path(out_dir, paste0("example_heatmap_zoom_", chr_show, "_", cell_name, ".png")), p_hm, width = 15, height = 6, dpi = 150)
  ggsave(file.path(out_dir, paste0("example_heatmap_diff_", chr_show, "_", cell_name, ".png")), p_diff, width = 8, height = 6, dpi = 150)
}
```

## 7) Embedding + scHiCTools Parity (Raw Only)

```{r embedding_eval}
if (!isTRUE(cfg$run_full)) {
  stop("Set cfg$run_full = TRUE to compute benchmark outputs.")
}

q_cov <- stats::quantile(cov_full$coverage, probs = cfg$embedding_eval_cov_q, na.rm = TRUE)
keep_eval <- cov_full$coverage >= q_cov[1] & cov_full$coverage <= q_cov[2]
cat("Embedding eval coverage filter:", round(q_cov[1], 4), "to", round(q_cov[2], 4),
    "| kept", sum(keep_eval), "of", length(keep_eval), "cells\n")

Y_raw <- Y_full[, keep_eval, drop = FALSE]
labels_eval <- labels_full[keep_eval]
cells_eval <- colnames(Y_raw)

infer_resolution_mb_from_path <- function(path) {
  b <- tolower(basename(path))
  m1 <- regexec("([0-9]+)m\\.scool$", b)
  g1 <- regmatches(b, m1)[[1]]
  if (length(g1) == 2) return(as.numeric(g1[2]))
  m2 <- regexec("([0-9]+)kb\\.scool$", b)
  g2 <- regmatches(b, m2)[[1]]
  if (length(g2) == 2) return(as.numeric(g2[2]) / 1000)
  1
}

infer_resolution_name <- function(path) {
  b <- tolower(basename(path))
  m1 <- regexec("([0-9]+)m\\.scool$", b)
  g1 <- regmatches(b, m1)[[1]]
  if (length(g1) == 2) return(paste0(g1[2], "M"))
  m2 <- regexec("([0-9]+)kb\\.scool$", b)
  g2 <- regmatches(b, m2)[[1]]
  if (length(g2) == 2) return(paste0(g2[2], "kb"))
  "1M"
}

strata_from_max_range_mb <- function(max_range_mb, resolution_mb) {
  if (!is.finite(max_range_mb)) return(as.integer(1e6))
  n <- floor(max_range_mb / resolution_mb) + 1
  as.integer(max(1, n))
}

resolution_mb <- infer_resolution_mb_from_path(matrix_file)
resolution_name <- infer_resolution_name(matrix_file)
cat("Inferred matrix resolution (Mb):", resolution_mb, "\n")

subset_embedding_chromosomes <- function(Y, mode = c("score_default", "autosomes_only", "all_chr")) {
  mode <- match.arg(mode)
  rn <- rownames(Y)
  if (is.null(rn) || !any(grepl("::", rn, fixed = TRUE))) return(Y)
  chr <- sub("::.*$", "", rn)
  if (identical(mode, "score_default")) {
    keep <- !(chr %in% c("chrY", "chrM", "MT", "M"))
  } else if (identical(mode, "autosomes_only")) {
    keep <- !(chr %in% c("chrX", "chrY", "chrM", "MT", "M"))
  } else {
    keep <- !(chr %in% c("chrM", "MT", "M"))
  }
  Y[keep, , drop = FALSE]
}

pipeline_to_settings <- function(pipe, resolution_mb, default_n = 10L) {
  out <- list(pipe = pipe, range_label = "default", n_strata = as.integer(default_n))
  if (grepl("_n[0-9]+$", pipe)) {
    n <- suppressWarnings(as.integer(sub("^.*_n([0-9]+)$", "\\1", pipe)))
    out$range_label <- paste0("n", n)
    out$n_strata <- n
    return(out)
  }
  if (grepl("_[0-9]+M$", pipe)) {
    m <- suppressWarnings(as.integer(sub("^.*_([0-9]+)M$", "\\1", pipe)))
    out$range_label <- paste0(m, "M")
    out$n_strata <- strata_from_max_range_mb(m, resolution_mb)
    return(out)
  }
  out
}

compute_sklearn_ari <- function(emb_xy, labels, k, nstart = 50L, seed = 12345L) {
  if (!requireNamespace("reticulate", quietly = TRUE)) return(NA_real_)
  sk <- tryCatch(reticulate::import("sklearn", delay_load = FALSE), error = function(e) NULL)
  if (is.null(sk)) return(NA_real_)
  km <- sk$cluster$KMeans(
    n_clusters = as.integer(k),
    n_init = as.integer(nstart),
    random_state = as.integer(seed)
  )
  pred <- km$fit_predict(as.matrix(emb_xy))
  ari <- sk$metrics$adjusted_rand_score(as.integer(factor(labels)), pred)
  as.numeric(ari)
}

chrom_modes <- tibble::tribble(
  ~chrom_mode, ~mode_key,
  "A_score_default", "score_default",
  "B_autosomes_only", "autosomes_only",
  "C_all_chr", "all_chr"
)

run_r_one <- function(Yin, labels, chrom_mode, pipe) {
  st <- pipeline_to_settings(pipe, resolution_mb, cfg$innerproduct_n_strata)
  cfg_i <- cfg
  cfg_i$innerproduct_n_strata <- as.integer(st$n_strata)

  out <- embed_and_score(
    Y_feature_by_cell = Yin,
    labels = labels,
    method_name = "Raw",
    pipeline = "paper_innerproduct_mds",
    cfg = cfg_i,
    coverage_vec = as.numeric(cov_full$coverage[keep_eval])
  )

  k <- length(unique(labels))
  ari_sklearn <- compute_sklearn_ari(out[, c("X1", "X2")], out$label, k, cfg$k_nstart, cfg$seed)

  out %>%
    dplyr::mutate(
      branch = "R",
      chrom_mode = chrom_mode,
      pipeline = pipe,
      range_label = st$range_label,
      n_strata = as.integer(st$n_strata),
      ari_r_kmeans = as.numeric(ari),
      ari_sklearn_kmeans = ari_sklearn
    )
}

emb_all <- dplyr::bind_rows(lapply(seq_len(nrow(chrom_modes)), function(i) {
  mode_name <- chrom_modes$chrom_mode[[i]]
  mode_key <- chrom_modes$mode_key[[i]]
  Y_mode <- subset_embedding_chromosomes(Y_raw, mode = mode_key)
  cat("R branch", mode_name, ": features=", nrow(Y_mode), " cells=", ncol(Y_mode), "\n")
  dplyr::bind_rows(lapply(cfg$embedding_pipelines, function(pipe) {
    set.seed(cfg$seed)
    run_r_one(Y_mode, labels_eval, mode_name, pipe)
  }))
}))

r_score_tbl <- emb_all %>%
  dplyr::distinct(branch, chrom_mode, pipeline, range_label, n_strata, ari_r_kmeans, ari_sklearn_kmeans) %>%
  dplyr::arrange(chrom_mode, n_strata, pipeline)

print(r_score_tbl)
```

```{r schictools_reference}
schic_ref_tbl <- tibble::tibble(
  chrom_mode = character(),
  pipeline = character(),
  range_label = character(),
  n_strata = integer(),
  ari_schictools_kmeans = double(),
  status = character()
)
schic_emb_all <- tibble::tibble()

if (isTRUE(cfg$run_schictools_reference)) {
  dir.create(cfg$schictools_out_dir, recursive = TRUE, showWarnings = FALSE)

  resolve_python_bin <- function(candidate) {
    if (!is.null(candidate) && nzchar(candidate) && file.exists(candidate)) {
      return(normalizePath(candidate, winslash = "/", mustWork = TRUE))
    }
    py <- if (!is.null(candidate) && nzchar(candidate)) Sys.which(candidate) else ""
    if (is.null(py) || !nzchar(py)) py <- Sys.which("python")
    if (nzchar(py) && grepl("WindowsApps[/\\\\]python\\.exe$", py, ignore.case = TRUE)) return("")
    py
  }

  run_python_script <- function(py, code_lines, args = character()) {
    if (!nzchar(py)) return(list(status = 1L, out = "python_not_found"))
    tf <- tempfile(fileext = ".py")
    writeLines(code_lines, tf)
    out <- suppressWarnings(system2(py, c(tf, args), stdout = TRUE, stderr = TRUE))
    st <- attr(out, "status")
    if (is.null(st)) st <- 0L
    list(status = as.integer(st), out = out)
  }

  python_has_module <- function(py, mod) {
    res <- run_python_script(py, c(
      "import importlib.util, sys",
      paste0("sys.exit(0 if importlib.util.find_spec('", mod, "') is not None else 1)")
    ))
    isTRUE(res$status == 0L)
  }

  py_bin <- resolve_python_bin(cfg$schictools_python)
  use_python <- nzchar(py_bin)
  precheck <- "ok"

  if (use_python) {
    py_info <- run_python_script(py_bin, c(
      "import sys, site",
      "print(sys.executable)",
      "print(sys.version)",
      "print(site.ENABLE_USER_SITE)"
    ))$out
    cat("scHiCTools python precheck info:\n", paste(py_info, collapse = "\n"), "\n", sep = "")
    required_mods <- c("numpy", "pandas", "scipy", "cooler", "scHiCTools")
    missing_mods <- required_mods[!vapply(required_mods, function(m) python_has_module(py_bin, m), logical(1))]
    if (length(missing_mods) > 0) {
      use_python <- FALSE
      precheck <- paste0("python_missing_modules_", paste(missing_mods, collapse = "_"))
      cat("Missing python modules for scHiCTools branch: ", paste(missing_mods, collapse = ", "), "\n", sep = "")
      cat("Fix with: ", py_bin, " -m pip install ", paste(missing_mods, collapse = " "), "\n", sep = "")
    }
  } else {
    precheck <- "python_not_found_or_windowsapps_alias"
    cat("scHiCTools precheck: usable Python not found.\n")
  }

  settings_tbl <- dplyr::bind_rows(lapply(cfg$embedding_pipelines, function(pipe) {
    st <- pipeline_to_settings(pipe, resolution_mb, cfg$innerproduct_n_strata)
    tibble::tibble(
      pipeline = pipe,
      range_label = st$range_label,
      n_strata = as.integer(st$n_strata)
    )
  }))

  if (!use_python) {
    schic_ref_tbl <- tidyr::crossing(
      chrom_mode = chrom_modes$chrom_mode,
      settings_tbl
    ) %>%
      dplyr::mutate(
        ari_schictools_kmeans = NA_real_,
        status = precheck
      ) %>%
      dplyr::select(chrom_mode, pipeline, range_label, n_strata, ari_schictools_kmeans, status)
  } else {
    cells_file <- file.path(cfg$schictools_out_dir, "cells_eval.txt")
    writeLines(cells_eval, cells_file)
    label_map <- tibble::tibble(cell = cells_eval, label = labels_eval)

    py_code <- c(
      "import argparse, os, json",
      "import numpy as np, pandas as pd",
      "import cooler",
      "from scipy.sparse import save_npz",
      "from scHiCTools import scHiCs",
      "",
      "p = argparse.ArgumentParser()",
      "p.add_argument('--scool', required=True)",
      "p.add_argument('--cells_file', required=True)",
      "p.add_argument('--resolution', type=int, required=True)",
      "p.add_argument('--out_dir', required=True)",
      "p.add_argument('--n_list', required=True)",
      "p.add_argument('--chrom_mode', required=True)",
      "args = p.parse_args()",
      "",
      "with open(args.cells_file, 'r', encoding='utf-8') as f:",
      "    cells = [x.strip() for x in f if x.strip()]",
      "n_list = [int(x) for x in args.n_list.split(',') if x.strip()]",
      "n_list = sorted(set(n_list))",
      "os.makedirs(args.out_dir, exist_ok=True)",
      "network_root = os.path.join(args.out_dir, 'network')",
      "os.makedirs(network_root, exist_ok=True)",
      "network = []",
      "chr_lengths = {}",
      "common_bins = {}",
      "cell_keep = {}",
      "# pass 1: determine common (minimum) bins per chromosome across cells",
      "for cell in cells:",
      "    cu = f\"{args.scool}::/cells/{cell}\"",
      "    c = cooler.Cooler(cu)",
      "    if args.chrom_mode == 'A_score_default':",
      "        keep = [ch for ch in c.chromnames if ch not in ('chrY', 'chrM', 'chrMT', 'Y', 'M', 'MT')]",
      "    elif args.chrom_mode == 'B_autosomes_only':",
      "        keep = [ch for ch in c.chromnames if ch not in ('chrX', 'chrY', 'chrM', 'chrMT', 'X', 'Y', 'M', 'MT')]",
      "    else:",
      "        keep = [ch for ch in c.chromnames if ch not in ('chrM', 'chrMT', 'M', 'MT')]",
      "    cell_keep[cell] = keep",
      "    for ch in keep:",
      "        try:",
      "            e0, e1 = c.extent(ch)",
      "            nb = int(e1 - e0)",
      "        except Exception:",
      "            continue",
      "        if nb <= 0:",
      "            continue",
      "        if (ch not in common_bins) or (nb < common_bins[ch]):",
      "            common_bins[ch] = nb",
      "",
      "if len(common_bins) == 0:",
      "    raise RuntimeError('No valid chromosomes with non-zero bins found')",
      "",
      "# pass 2: write fixed-shape per-chromosome matrices",
      "for cell in cells:",
      "    cu = f\"{args.scool}::/cells/{cell}\"",
      "    c = cooler.Cooler(cu)",
      "    keep = cell_keep.get(cell, [])",
      "    cell_dir = os.path.join(network_root, cell)",
      "    os.makedirs(cell_dir, exist_ok=True)",
      "    for ch in keep:",
      "        nb = int(common_bins.get(ch, 0))",
      "        if nb <= 0:",
      "            continue",
      "        try:",
      "            e0, e1 = c.extent(ch)",
      "            if (e1 - e0) <= 0:",
      "                continue",
      "            mat = c.matrix(balance=False, sparse=True).fetch(ch).tocsr()",
      "            mat = mat[:nb, :nb].tocoo()",
      "        except Exception:",
      "            continue",
      "        save_npz(os.path.join(cell_dir, ch), mat)",
      "        if ch not in chr_lengths:",
      "            chr_lengths[ch] = int(nb * args.resolution)",
      "    network.append(cell_dir)",
      "",
      "if len(network) == 0:",
      "    raise RuntimeError('No cells prepared for scHiCTools run')",
      "",
      "for n in n_list:",
      "    loaded = scHiCs(network, reference_genome=chr_lengths, resolution=args.resolution, chromosomes='all',",
      "                   keep_n_strata=int(n), strata_offset=0, exclusive_strata=False, store_full_map=False,",
      "                   format='npz', operations=None, t=1, random_walk_ratio=1.0)",
      "    emb, _ = loaded.learn_embedding(dim=2, similarity_method='innerproduct', embedding_method='MDS',",
      "                                    aggregation='median', print_time=False, return_distance=True, n_strata=int(n))",
      "    out = pd.DataFrame({'cell': cells, 'X1': emb[:, 0], 'X2': emb[:, 1], 'n_strata': int(n)})",
      "    out.to_csv(os.path.join(args.out_dir, f'embedding_n{int(n)}.csv'), index=False)",
      "print('ok')"
    )

    run_mode <- function(chrom_mode) {
      mode_dir <- file.path(cfg$schictools_out_dir, chrom_mode)
      dir.create(mode_dir, recursive = TRUE, showWarnings = FALSE)
      n_vals <- sort(unique(settings_tbl$n_strata))
      py_args <- c(
        "--scool", matrix_file,
        "--cells_file", cells_file,
        "--resolution", as.character(as.integer(resolution_mb * 1e6)),
        "--out_dir", mode_dir,
        "--n_list", paste(n_vals, collapse = ","),
        "--chrom_mode", chrom_mode
      )
      out <- run_python_script(py_bin, py_code, args = py_args)
      st_txt <- if (isTRUE(out$status == 0L)) "ok" else paste0("schictools_run_failed_exit_", out$status)
      log_file <- file.path(mode_dir, "run.log")
      writeLines(out$out, con = log_file)

      one_tbl <- settings_tbl %>%
        dplyr::mutate(chrom_mode = chrom_mode, ari_schictools_kmeans = NA_real_, status = st_txt) %>%
        dplyr::select(chrom_mode, pipeline, range_label, n_strata, ari_schictools_kmeans, status)

      emb_mode <- tibble::tibble()
      if (identical(st_txt, "ok")) {
        emb_mode <- dplyr::bind_rows(lapply(cfg$embedding_pipelines, function(pipe) {
          st <- pipeline_to_settings(pipe, resolution_mb, cfg$innerproduct_n_strata)
          f <- file.path(mode_dir, paste0("embedding_n", as.integer(st$n_strata), ".csv"))
          if (!file.exists(f)) return(NULL)
          df <- suppressWarnings(readr::read_csv(f, show_col_types = FALSE, progress = FALSE))
          req <- c("cell", "X1", "X2")
          if (!all(req %in% names(df))) return(NULL)
          df <- df %>% dplyr::rename(cell = .data$cell)
          df <- df %>% dplyr::left_join(label_map, by = "cell")
          labs <- as.character(df$label)
          good <- !is.na(labs) & is.finite(df$X1) & is.finite(df$X2)
          ari_val <- NA_real_
          if (sum(good) >= 3) {
            km <- stats::kmeans(df[good, c("X1", "X2"), drop = FALSE], centers = length(unique(labs[good])), nstart = cfg$k_nstart)
            ari_val <- suppressWarnings(mclust::adjustedRandIndex(as.integer(as.factor(labs[good])), km$cluster))
          }
          one_tbl$ari_schictools_kmeans[one_tbl$pipeline == pipe] <<- ari_val
          df %>% dplyr::transmute(branch = "scHiCTools", chrom_mode = chrom_mode, pipeline = pipe, range_label = st$range_label, n_strata = as.integer(st$n_strata), X1, X2, label = .data$label, ari_schictools_kmeans = ari_val)
        }))
      }
      list(tbl = one_tbl, emb = emb_mode)
    }

    mode_results <- lapply(chrom_modes$chrom_mode, run_mode)
    schic_ref_tbl <- dplyr::bind_rows(lapply(mode_results, `[[`, "tbl"))
    schic_emb_all <- dplyr::bind_rows(lapply(mode_results, `[[`, "emb"))
  }
}

print(schic_ref_tbl)
```

## 8) Parity Tables + Plots

```{r parity_plots, fig.width=12, fig.height=6}
parity_tbl <- r_score_tbl %>%
  dplyr::left_join(
    schic_ref_tbl %>% dplyr::select(chrom_mode, pipeline, ari_schictools_kmeans, schictools_status = status),
    by = c("chrom_mode", "pipeline")
  ) %>%
  dplyr::mutate(
    chrom_mode_label = dplyr::recode(
      chrom_mode,
      "A_score_default" = "Autosome + chrX (SCORE default)",
      "B_autosomes_only" = "Autosome only",
      "C_all_chr" = "Autosome + chrX + chrY (all chr)"
    )
  ) %>%
  dplyr::arrange(chrom_mode, n_strata, pipeline)

print(parity_tbl)

curve_df <- parity_tbl %>%
  dplyr::mutate(
    range_ord = factor(range_label, levels = c("default", "1M", "2M", "4M", "6M", "10M", "15M", "20M", "30M", "50M", "n10", "n11", "n32"))
  ) %>%
  tidyr::pivot_longer(
    cols = c(ari_r_kmeans, ari_sklearn_kmeans, ari_schictools_kmeans),
    names_to = "metric",
    values_to = "ari"
  ) %>%
  dplyr::mutate(
    metric = dplyr::recode(
      metric,
      "ari_r_kmeans" = "R embedding + R k-means",
      "ari_sklearn_kmeans" = "R embedding + sklearn k-means",
      "ari_schictools_kmeans" = "scHiCTools embedding + R k-means"
    )
  )

p_curve <- ggplot(curve_df, aes(x = range_ord, y = ari, color = metric, group = metric)) +
  geom_line(linewidth = 0.8, na.rm = TRUE) +
  geom_point(size = 1.8, na.rm = TRUE) +
  facet_wrap(~chrom_mode_label, ncol = 1) +
  theme_minimal() +
  labs(
    title = "Flayamer InnerProduct Parity Curve (1M, Raw)",
    subtitle = "Embedding method + k-means method",
    x = "Distance / Strata setting",
    y = "ARI",
    color = "Metric"
  )
print(p_curve)
ggsave(file.path(out_dir, "innerproduct_parity_curve_1M_raw.png"), p_curve, width = 12, height = 7, dpi = 150)

plot_df <- dplyr::bind_rows(
  emb_all %>% dplyr::mutate(branch = "R"),
  schic_emb_all
) %>%
  dplyr::left_join(
    parity_tbl %>% dplyr::select(chrom_mode, chrom_mode_label, pipeline, ari_schictools_kmeans_ref = ari_schictools_kmeans),
    by = c("chrom_mode", "pipeline")
  ) %>%
  dplyr::mutate(
    ari_schictools_kmeans = dplyr::coalesce(.data$ari_schictools_kmeans, .data$ari_schictools_kmeans_ref)
  ) %>%
  dplyr::mutate(panel = paste0(
    branch, " | ", chrom_mode_label, "\nR=", round(ari_r_kmeans, 3),
    " | sklearn=", round(ari_sklearn_kmeans, 3),
    " | scHiCTools=", round(ari_schictools_kmeans, 3)
  ))

for (pipe in unique(plot_df$pipeline)) {
  p_one <- ggplot(plot_df %>% dplyr::filter(pipeline == pipe), aes(X1, X2, color = label)) +
    geom_point(size = 1, alpha = 0.9) +
    facet_wrap(~panel, ncol = 2, scales = "free") +
    theme_minimal() +
    labs(
      title = paste0("Flayamer InnerProduct Embedding: ", pipe),
      subtitle = paste0("n_cells=", sum(keep_eval), " | Raw only"),
      color = label_col
    )
  print(p_one)
  plot_path <- file.path(out_dir, paste0("embedding_parity_", pipe, ".png"))
  ggsave(plot_path, p_one, width = 12, height = 6, dpi = 150)
}
```

```{r}
curve_df
```


## 9) Analysis Summary

- Data and setup
  - Dataset: Flayamer oocyte/zygote, 1M resolution.
  - Raw-only InnerProduct parity run targeting k-means ARI.

- QC
  - Depth filter from metadata: depth >= 5000.
  - Chromosome-level filter: each autosome of length x Mb must have at least x cis-contacts.
  - Cell removed if any autosome fails.

- Branches compared
  - R branch: paper_innerproduct_mds (current R implementation).
  - scHiCTools branch: direct scHiCTools InnerProduct on the same .scool + filtered cell set.

- Distance/strata settings
  - default, 1M, 2M, 4M, 6M, 10M, 15M, 20M, 30M, 50M
  - explicit strata checks: n10, n11, n32

- Evaluation
  - ARI from R kmeans and sklearn kmeans on the same R embedding.
  - ARI from scHiCTools embedding + R kmeans on matched labels.
  - Parity curve and embedding plots annotated with all three ARI values.

